---
topic: "land"
title: "Trends in Timber Harvest in B.C."
output: envreportbc::print_ver
---


```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library(readr) #read CSV file
library(dplyr) #data prep
library(reshape2) #for restructuring data table
library(ggplot2)  #for plotting
library(extrafont) #Verdana font
library(envreportutils) #soe theme
library(rphylopic) #for conifer image, package available from GitHub only
library(curl) #required by rphylopic function
library(knitr) #for knitting to PDF

read_chunk("timber_harvest.R")
```

```{r pre}
```

Timber harvesting has direct implications for British Columbia's
environmental integrity and sustainable economic activity. Timber
harvesting has contributed substantially to B.C.′s economic development
and continues to provide the economic base for many rural communities.
Environmental integrity and sustainable economic activity depend on
maintaining a long-term balance between forest growth and timber
harvest.

-   **How much timber is harvested annually?** The annual volume of
    timber harvested in the province generally increased from
    1900---peaking at almost 90 million cubic metres in 1987. Since
    then, the annual harvest has fluctuated around an average of 77
    million cubic metres, of which 90% is regulated under the
    government-set Allowable Annual Cut.
-   **How does the actual harvest compare with the Allowable Annual
    Cut?** Generally the amount of timber harvested each year has been
    below the Allowable Annual Cut on public lands in British Columbia.
-   **What is the provincial timber supply forecast?** Predictable
    timber supplies are important to the provincial economy and to
    customers who rely on B.C.′s forest products. Timber supply is
    forecasted in response to changes in the forest and new information,
    forecasting methods, and assumptions. Allowable Annual Cuts are
    required by law to be determined at least every ten years, using the
    latest information to reduce the risks related to uncertainty.
    Uncertainties in timber supply forecasting include merchantability
    of tree species, operability, future management practices, the
    future timber production land base, wildfires and insect epidemics.
    In reality, most Allowable Annual Cut decisions are revisited every
    5 years to account for changes in timber supply and harvesting
    practices.

\vspace{30pt}

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
  
   \begingroup\centering{\textbf{What is the Allowable Annual Cut?}}
   
     \endgroup
     
    \vspace{1mm}
    
    \text{The Allowable Annual Cut is the maximum average level of timber harvest permitted}     \text{for forest management areas---it represents a harvest level that balances}             \text{environmental, economic and social considerations.}
 
 \vspace{3mm}    
 
    \text{Since 1949, the Chief Forester or their delegate has set Allowable Annual Cuts to}     \text{regulate harvest levels on public land and some private land.}
    
    
  \end{minipage}
}
\end{center}

\newpage

## How Much Timber Is Harvested Annually?

\vspace{20pt}

```{r harvest, fig.height = 6, fig.width = 11}
```

\vspace{20pt}

-   The annual timber harvest from all public and private land increased
    10-fold since the 1900s, and generally levelled off in the 1990s.
    Over the last 10 years, the average total timber harvest was 74
    million cubic metres per year, of which 67 million cubic metres per
    year (90%) was from forests where harvest levels were regulated by
    Allowable Annual Cuts. The remaining 7 million cubic metres of
    timber harvested per year (10% of the total harvest) was from lands
    with no government-set Allowable Annual Cuts.

\newpage

## How Does the Actual Timber Harvest Compare with the Allowable Annual Cut?

\vspace{20pt}

```{r aac, fig.height = 8, fig.width = 12}
```

\vspace{20pt}

-   Areas regulated by government-set Allowable Annual Cuts account
    for 90% of the total harvest volume.
-   Actual timber harvest can be compared to the average Allowable
    Annual Cut. Over the last 10 years, the amount of timber harvested
    annually on public land has been below the Allowable Annual Cut. The
    average Allowable Annual Cut was 83 million cubic metres per year
    and the average harvest on these areas was 67 million cubic metres
    per year.
-   In the last 10 years, the Allowable Annual Cut was increased in
    mountain pine beetle impacted areas of the province. Beginning in
    2006, a downturn in market conditions resulted in decreasing actual
    harvest despite the increases in Allowable Annual Cut.

\newpage

## What Is the Provincial Timber Supply Forecast?

\vspace{20pt}

```{r forecast, fig.height = 6, fig.width = 12}
```

\vspace{20pt}

-   Until relatively recently, B.C. was forecast to have a stable mid
    and long-term timber supply of about 70 million cubic meters per
    year. Recent analysis projects decreases in timber supply to about
    58 million cubic metres per year by 2025---due to mortality caused
    by the mountain pine beetle epidemic. The forecasted timber supply
    returns to approximately 65--70 million cubic metres per year
    by 2075.
-   Detailed timber supply forecasts and their related assumptions are
    [publicly available for forest management
    areas](http://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598).

\newpage

## Methods
Timber harvest data reported here is from public and private land that
is regulated by the [government-set Allowable Annual
Cuts](https://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598).
Allowable Annual Cuts are as of December 31st of the year listed---the
starting year of fiscal years. Harvest data are by calendar year for
1945--1979 and fiscal year (April 1st to March 31st) for 1980 onward.
This indicator is a composite from the [Timber Supply
Review](https://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598)
process and the [Harvest Billing
System](https://www2.gov.bc.ca/gov/content?id=4E1093E8C2D9472EA82B2CB00A6C59FD).
The provincial Timber Supply Forecast is a summary of the management
unit harvest forecast data.

## References and Other Useful Links
* [Learn more about Forest Stewardship in
    B.C.](https://www2.gov.bc.ca/gov/content?id=EE19793D89C149E6A16CD6B71AF71FF0)
* [Timber Supply Review & Allowable Annual
    Cut](http://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598)
* [Harvest Billing
    System](https://www2.gov.bc.ca/gov/content?id=4E1093E8C2D9472EA82B2CB00A6C59FD)


## Data

\*By accessing these datasets, you agree to the licence associated with each file, as indicated in parentheses below.

* [Indicator data: B.C. Timber Harvest (1910-2015)]() (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
* [Indicator data: B.C. Timber Supply Forecast (2010-2100)]() (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (December 2017):
[http://www.env.gov.bc.ca/soe/indicators/land/timber-harvest.html](http://www.env.gov.bc.ca/soe/indicators/land/timber-harvest.html)

Email correspondence to: envreportbc@gov.bc.ca
