---
topic: "land"
title: "Trends in Timber Harvest in B.C."
output: envreportutils.internal::print_ver
---
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library(readr) #read CSV file
library(dplyr) #data prep
library(reshape2) #for restructuring data table
library(ggplot2)  #for plotting
library(extrafont) #Verdana font
library(envreportutils) #soe theme
library(rphylopic) #for conifer image, package available from GitHub only
library(curl) #required by rphylopic function
library(knitr) #for knitting to PDF
library(bcdata)
library(here)

read_chunk("timber_harvest.R")
```

```{r pre}
```

Timber harvesting has direct implications for British Columbia's
environmental integrity and sustainable economic activity. Timber
harvesting has contributed substantially to B.C.'s economic development
and continues to provide the economic base for many rural communities.
Environmental integrity and sustainable economic activity depend on
maintaining a long-term balance between forest growth and timber
harvest.

-   **How much timber is harvested annually?** The annual volume of timber harvested in the province generally increased since 1900, to a peak of almost 90 million cubic metres in 1987. From 1988 to 2013, the annual harvest fluctuated around an average of 75 million cubic metres.  Since 2014, the annual harvest has declined from 71 to 39 million cubic metres.
-   **How does the actual harvest compare with the Allowable Annual Cut?** Generally the amount of timber harvested each year has been
    below the Allowable Annual Cut on public lands in British Columbia.
-   **What is the provincial timber supply forecast?** Predictable timber supplies are important to the provincial economy and to customers who rely on B.C.′s forest products. Timber supply is forecasted in response to changes in the forest and new information, forecasting methods, and assumptions. Allowable Annual Cuts are required by law to be determined at least every ten years, using the latest information to reduce the risks related to uncertainty. Uncertainties in timber supply forecasting include merchantability of tree species, operability, future management practices, the future timber production land base, wildfires and insect epidemics.

\vspace{30pt}

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
  
   \begingroup\centering{\textbf{What is the Allowable Annual Cut?}}
   
     \endgroup
     
    \vspace{1mm}
    
    \text{The Allowable Annual Cut is the maximum average level of timber harvest permitted}     \text{for forest management areas---it represents a harvest level that balances}             \text{environmental, economic and social considerations.}
 
  \end{minipage}
}
\end{center}

\newpage

## How Much Timber Is Harvested Annually?

- The annual volume of timber harvested in the province generally increased from 1900 to a peak of almost 90 million cubic metres in 1987. From 1988 to 2013, the annual harvest fluctuated around an average of 75 million cubic metres.  Since 2014, the annual harvest has declined from 71 to 39 million cubic metres. 
- Since 1949, the Chief Forester or their delegate has set Allowable Annual Cuts to     regulate harvest levels on public land and some private land. Harvest without regulation by an Allowable Annual Cut occurs on private lands under the *Private Managed Forest Land Act*.
-	Since 1987, the average total timber harvest was 72 million cubic metres per year, of which 64 million cubic metres per year (89%) was from forests where harvest levels were regulated by Allowable Annual Cuts. The remaining 8 million cubic metres of timber harvested per year (11% of the total harvest) was from lands with no government-set Allowable Annual Cuts.

```{r harvest, fig.height = 6, fig.width = 11}
```

\vspace{20pt}

\newpage

## How Does the Actual Timber Harvest Compare with the Allowable Annual Cut?

- 	Over the last 20 years, the amount of timber harvested annually on public land has been 22% below the Allowable Annual Cut. The average Allowable Annual Cut during this period was 77 million cubic metres while the average harvest was 60 million cubic metres per year.
-	From 2001 to 2014, the Allowable Annual Cut was increased in mountain pine beetle impacted areas of the interior. 
- From 2006 to 2009, a downturn in market conditions resulted in decreasing actual harvest despite the increases in Allowable Annual Cut.

\vspace{20pt}

```{r aac, fig.height = 6, fig.width = 11}
```

\newpage

## What Is the Provincial Timber Supply Forecast?

-   Until the early 2000’s, B.C. was forecasted to have a stable mid- and long-term timber supply of about 70 million cubic metres per year.  Recent Timber Supply Review Allowable Annual Cut decisions have reduced the forecasted mid- and long-term timber supply to about 57 and 65 million cubic metres respectively. The main causes for this reduction are mortality due to the mountain pine beetle, mortality due to catastrophic wildfires, and an increasing amount of [conservation areas](https://www.env.gov.bc.ca/soe/indicators/land/land-designations.html)  where harvesting is reduced or not permitted.
- Detailed timber supply forecasts and their related assumptions are [publicly available for forest management areas](http://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598).

\vspace{20pt}

```{r forecast, fig.height = 6, fig.width = 11}
```

\newpage

## Methods

Timber harvest data reported here is from public and private land that
is regulated by the [government-set Allowable Annual
Cuts](https://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598).
Allowable Annual Cuts are as of December 31st of the year listed---the
starting year of fiscal years. Harvest data are by calendar year for
1945--1979 and fiscal year (April 1st to March 31st) for 1980 onward.
This indicator is a composite from the [Timber Supply
Review](https://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598)
process and the [Harvest Billing
System](https://www2.gov.bc.ca/gov/content?id=4E1093E8C2D9472EA82B2CB00A6C59FD).
The provincial Timber Supply Forecast is a summary of the management
unit harvest forecast data. 

For more information on this indicator or on Timber Harvest in British Columbia contact the Forest Analysis and Inventory Branch at [Forests.ForestAnalysisBranchOffice@gov.bc.ca](Forests.ForestAnalysisBranchOffice@gov.bc.ca).

The [**R** code](http://www.r-project.org) for creating the graphs presented on this page is available on [GitHub](https://github.com/bcgov/forest-management-indicators).

## References and Other Useful Links
* [Learn more about Forest Stewardship in
    B.C.](https://www2.gov.bc.ca/gov/content?id=EE19793D89C149E6A16CD6B71AF71FF0)
* [Timber Supply Review & Allowable Annual
    Cut](http://www2.gov.bc.ca/gov/content?id=EAEB71451EB949DFB14A9E1957CF9598)
* [Harvest Billing
    System](https://www2.gov.bc.ca/gov/content?id=4E1093E8C2D9472EA82B2CB00A6C59FD)

## Data

\*By accessing these datasets, you agree to the licence associated with each file, as indicated in parentheses below.

* [Indicator Summary Data: Change in Timber Harvest in B.C. (1910-2023)](https://catalogue.data.gov.bc.ca/dataset/18754165-1daa-42ea-8c43-fef0e8cf4598) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
* [Indicator Summary Data: B.C. Timber Supply Forecast (2010-2210)](https://catalogue.data.gov.bc.ca/dataset/18754165-1daa-42ea-8c43-fef0e8cf4598) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (June 2025):
[http://www.env.gov.bc.ca/soe/indicators/land/timber-harvest.html](http://www.env.gov.bc.ca/soe/indicators/land/timber-harvest.html)

Email correspondence to: envreportbc@gov.bc.ca
