---
topic: "land"
title: "Trends in Silviculture in B.C. (1987-2015)"
output: envreportbc::print_ver
---


```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
require(knitr)
library(dplyr)
library(reshape2)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(envreportutils)
library(magrittr)
library(extrafont)

opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=5)
read_chunk("../silviculturegraphs.R")

options(stringsAsFactors=FALSE)

################################################################################
## Read in the data files
################################################################################

# Silvicultural systems
silsystems <- read.csv("Z:/forests/silviculture/2017/silviculture_systems.csv",
                        strip.white=TRUE)

# Disturbances and Reforestation
dist.refor <- read.csv("Z:/forests/silviculture/2017/disturbance_and_reforestation.csv",
                       strip.white=TRUE)

# Silvicultural Treatments
treatments <- read.csv("Z:/forests/silviculture/2017/silviculture_treatments.csv",
                       strip.white=TRUE)

# Gains
gains <- read.csv("Z:/forests/silviculture/2017/timber_volume_gains.csv",
                  strip.white=TRUE)

# Set chart font to "" so uses pdf font
chart_font_web <- ""
```

```{r pre}

```

Managing and conserving the province’s Crown forest and range resources provides environmental, social and economic benefits to all British Columbians. Silviculture---one of the primary means to enhance the social and economic benefits from our forest resource---entails the manipulation of forest and woodland vegetation in stands and on landscapes to meet the diverse needs and values of landowners and society on a sustainable basis. This indicator investigates the trends in four measures of silviculture from 1987 to 2015.


-   [Silviculture](#gloss) is the practice of growing and cultivating trees in
    forests to meet certain objectives, such as timber production,
    biodiversity, wildlife habitat or visual quality.

-   Silviculture activities occur within a silviculural system, which is
    a planned cycle of forest or tree harvest, regeneration and tending
    over time. Silvicultural systems used in British Columbia are
    [clearcutting](#gloss), partial cutting ([seed tree](#gloss), [shelterwood](#gloss),
    [selection](#gloss)), [clearcutting with reserves](#gloss) and 
    [variable retention](#gloss). Each name reflects the type of forest structure
    remaining after the initial harvest.

-   Changes in silvicultural practices over the past 40 years have
    increased reforestation, increased the volume and value of future
    timber supplies.

-   Silvicultural choices are often influenced by natural disturbances,
    public expectations and market demands.

```{r silviculture, fig.height=4.6}
```

-   Since 1970, the area of Crown forest harvested annually has ranged
    from 120,000 hectares to 240,000 hectares, with an average of
    roughly 175,000 hectares each year.

-   From 1970-1996, clearcutting systems were applied on 87% of the area
    harvested on public land; and partial cutting (seed tree,
    shelterwood, and Selection ) systems on 13%. By 2012/13 clearcutting
    with reserves and clearcutting accounted for 96% of the area
    harvested on public land with variable retention and partial cutting
    systems accounting for 4%.

```{r dist-refor}
```

-   Silvicultural investments can accelerate reforestation and thereby
    increase timber supplies and restore [ecological services](#gloss) sooner.

-   In the 1970s and 1980s on Crown forest land, the area disturbed
    exceeded the area reforested. As a result, the area deemed 
    [not satisfactorily restocked (NSR)](#gloss) after timber harvesting and other
    disturbances increased over several decades.

-   This gap was closed during the 1980s with increasing investments in
    [site preparation](#gloss), research that led to better planting methods,
    [planting](#gloss), [backlog planting](#gloss) and [brushing](#gloss) to ensure
    prompt restocking and the growth of desired trees.

-   In 1987, explicit reforestation obligations on public land were
    introduced. Since 1987, holders of harvesting rights are required to
    reforest the areas they harvest. This led to planting of a greater
    proportion of current harvest areas.

-   In recent years harvesting and reforestation are in step with each
    other. Since 2005, over 76 million seedlings have been planted on
    over 56,000ha of Mountain Pine Beetle and wildfire impacted land not
    currently under legal reforestation obligations.

```{r treat}
```


-   Silvicultural treatments are investments in future timber production
    and environmental benefits from forests. Silvicultural treatments
    can also increase timber quantity and quality, manage forest health
    and fire risks, and improve specific habitats, water management and
    visual landscapes. Silvicultural treatments also create employment
    opportunities for First Nations and others affected by changes in
    the forest industry.

-   Between 1976 and 2012, investments in [incremental silviculture](#gloss) to 
    improve the growth and quality of future crop trees included 
    [fertilizing](#gloss) (394,000 hectares), [pruning](#gloss) (57,000 hectares), 
    [spacing](#gloss) (678,000 hectares) and the use of [select seed](#gloss) for 
    planting (1,240,000 hectares). Over this period, silviculture treatments 
    totalled approximately 2.4 million hectares.

-   Public funding of fertilization, pruning, and spacing has been
    significantly reduced in recent years, however, funding is
    continuing for research, production, storage and monitoring of
    select seed use on public land. Fertilization has seen significant 
    increases since the inception of the [Forests for Tomorrow](#gloss) 
    program in 2004.  

-   By increasing the rate of tree growth, use of select seed can
    increase future timber volume, reduce constraints on harvesting
    adjacent areas and reduce the need for, and cost of, brushing.

```{r gains}
```

-   Investments in silviculture depend on expected gains in ecological
    services, volume and return on investment.

-   Compared with [natural regeneration](#gloss), planting increases
    harvestable volume by about 15%, and about 25% with the use of
    select seed.

-   Cumulative volume gains 65 years after making investments in incremental 
    silviculture since 1976 are estimated at 65 million cubic metres (m^3^), 
    based on fertilization (10 million m^3^), Spacing (11 million m^3^), 
    select seed (33 million m^3^), and reforestation through the Forests for 
    Tomorrow program (11 million m^3^). 

-   Other gains from silvicultural treatments include increased short
    and [mid-term timber supply](#gloss) through spacing and fertilization,
    accelerated development of mature or old growth forest
    characteristics where needed for wildlife or biodiversity, higher
    wood quality through pruning, more pleasing visual landscapes and
    planting to make up the long-term timber supply.


## References and other useful links
-   Ministry of Forests, Lands and Natural Resource Operations Annual
    Reports: <http://www.for.gov.bc.ca/mof/annualreports.htm>

-   Reporting Silviculture Updates and Land Status Tracking System
    (RESULTS): <http://www.for.gov.bc.ca/his/results/>

-   Silviculture Program Page:
    <http://www.for.gov.bc.ca/hfp/silviculture/index.htm>

-   Seed Planning and Registry Application (SPAR):
    <http://www.for.gov.bc.ca/HTI/spar/>
    
-   Forests for Tomorrow (FFT) reforestation program: 
    <http://www.for.gov.bc.ca/hcp/fia/landbase/fft/>


## Data
\*By accessing these datasets, you agree to the license associated with
each file, as indicated below.

-   Chart data: Silviculture Systems <http://catalogue.data.gov.bc.ca/dataset/forest-area-harvested-by-silviculture-systems>
    -   License: B.C. OGL: <http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf>

-   Chart data: Disturbances and Reforestation <http://catalogue.data.gov.bc.ca/dataset/trends-in-forest-disturbances-and-reforestation>
    -   License: B.C. OGL: <http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf>

-   Chart data: Silvicultural Treatments <http://catalogue.data.gov.bc.ca/dataset/trends-in-silvicultural-treatments-in-b-c>
    -   License: B.C. OGL: <http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf>

-   Chart data: Timber Volume Gains <http://catalogue.data.gov.bc.ca/dataset/expected-timber-volume-gains-from-silvicultural-treatments>
    -   License: B.C. OGL: <http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf>


## Glossary {#gloss}

Silviculture
:   The art and science of controlling the
    establishment, growth, composition, health and quality of forests
    and woodlands. Silviculture entails the manipulation of forest and
    woodland vegetation in stands and on landscapes to meet the diverse
    needs and values of landowners and society on a sustainable basis.

Clearcutting
:   The process of removing all trees, large and small,
    in a stand in one cutting operation. Clear-cutting with reserves is
    the removal of the majority of the trees saving some within or
    outside the cut boundary for other purposes such as wildlife
    habitat.

Seed tree
:   Trees left after reproduction cutting to provide seeds
    for natural regeneration. Trees are chosen for desired species,
    being healthy and and having good growth form.

Shelterwood
:   Harvesting trees in two or more cuttings that
    creates conditions where a new generation of seedlings can be
    established under the remaining timber without planting. This system
    is implemented in forests with species that can establish themselves
    in partial shade under the remaining mature trees.

Selection
:   Harvesting trees in a way that moves a forest stand
    towards an uneven-aged condition.

Clearcutting with reserves
:   The removal of the majority of the
    trees saving some within or outside the cut boundary for other
    purposes such as wildlife habitat.

Variable retention
:   Retaining forest structural elements for at
    least one rotation in order to preserve environmental values.

Ecosystem services
:   benefits to humankind and wildlife such as
    the production of food, provisioning of clean water, decomposition
    of wastes, control of micro-climate, regulating nutrient cycles,
    pollination, spiritual and recreational sites.

Not satisfactorily restocked (NSR)
:   productive forest land that
    has been denuded and has failed, partially or completely, to
    regenerate either naturally or by planting or seeding to the
    specified or desired free growing standards for the site.

Site preparation
:   any action, related to reforestation, to create
    an environment favourable for survival of suitable trees during the
    first growing season. It may alter the ground cover, soil or
    microsite conditions, using biological, mechanical, or manual
    clearing, prescribed burns, herbicides, or a combination of methods.
    Both natural regeneration and planting may be improved through site
    preparation.

Planting
:   establishing a forest by setting out seedlings,
    transplants or cuttings in an area.

Back-log planting
:   planting that is overdue. In general, planting
    is considered backlog planting if more than 7 years have elapsed
    since a site was cleared (by harvesting, fire, insects or disease)
    in the Interior, and more than 3 years have elapsed on the Coast of
    British Columbia.

Brushing
:   a silvicultural activity done by chemical, manual,
    grazing, or mechanical means to control competing forest vegetation
    and reduce competition for space, light, moisture, and nutrients
    with crop trees or seedlings.

Incremental Silviculture
:   Application of cultural measures which will allow an increase in the 
    value or volume of the cut over and above the establishment of a fully 
    stocked stand after harvest.

Fertilizing
:   the addition of fertilizer to promote tree growth on
    sites deficient in one or more soil nutrients. Also used to improve
    the vigor of crop trees following juvenile spacing or commercial
    thinning.
    
Forests for Tomorrow
:   A silviculture program, established by the B.C. Provincial Government 
    in March 2005 to respond to catastrophic wildfires and the mountain pine 
    beetle epidemic.

Pruning
:   the manual removal, close to or flush with the stem, of
    side branches, live or dead, and of multiple leaders from standing,
    generally plantation-grown trees. Pruning is carried out to improve
    the market value of the final wood product by producing knot-free
    wood for the improvement of the tree or its timber.

Spacing
:   the removal of undesirable trees within a young stand to
    control stocking, to maintain or improve growth, to increase wood
    quality and value, or to achieve other resource management
    objectives.

Select seed
:   seed that exhibits a higher level of one or more
    desired genetic traits (such as growth rate, form, wood density, and
    resistance to insects and disease) than wild seed collected from an
    average natural stand. This includes seed from tested parents
    growing in seed orchards and seed collected from natural stand
    superior provenances. Vegetative material for propagation, from
    production facilities using tested parents and from superior
    provenances, is included in the term select seed.

Natural regeneration
:   a change in forest structure and
    composition caused by fire, insects, wind, landslides and other
    natural processes.

Mid-term timber supply
:   the mid-term timber supply is a term that
    refers to that portion of the timber inventory that would be
    available for harvest within the middle of the normal management
    cycle.


*****
# METHODS - Trends in Silviculture in B.C. (1970-2012)

## Silviculture Systems:  
Detailed silviculture records are stored in the Reporting Silviculture
Updates and Land Status Tracking System (RESULTS;
<http://www.for.gov.bc.ca/his/results/>). Reported disturbances and
reforestation on public land are published in Ministry of Forests, Lands
and Natural Resources Operations annual reports (since 1989/90).

Data was taken from Ministry of Forests Lands and Natural Resource
Operations annual reports - Area Harvested by Silvicultural Systems on
Crown Land, by Forest Region (table 7.3) available here:
<http://www.for.gov.bc.ca/hfp/silviculture/statistics/statistics.htm>.
Private land accounts for 10% of the total area harvested in the
province between 1983/84 and 2008/09. Earlier data are not available.

Fiscal years were rounded down to calendar years in order to standardize
data display (e.g., 1980/1981 was considered to be 1980).

For the purposes of data rollup, partial cutting includes coppice,
intermediate cut, patch cut, seed tree, selection, shelterwood and
unspecified.

The areas for 1970 are estimated using the ratio of volumes harvested in
1970 and 1971, applied to the areas harvested in 1971.

## Disturbances and Reforestation:
Detailed silviculture records are stored in the Reporting Silviculture
Updates and Land Status Tracking System (RESULTS;
<http://www.for.gov.bc.ca/his/results/>).

Data was taken from Ministry of Forests Lands and Natural Resource
Operations annual reports - Changes in the Not Satisfactorily Restocked
(NSR) Crown Land (pdf) available here:
<http://www.for.gov.bc.ca/hfp/silviculture/statistics/statistics.htm>.

Fiscal years were rounded down to calendar years in order to standardize
data display (e.g., 1980/1981 was considered to be 1980).

Disturbances or 'Additions to NSR' are areas that become not
satisfactorily restocked (NSR) with timber due to harvesting, fire,
insects, diseases and other causes. Data for 1989/90 onwards are based
on surveyed and reported changes to NSR areas of public (Crown) land,
from ministry annual reports. For earlier years, estimates are based on
areas harvested and planted, and estimated rates of losses to fires,
insects, diseases and restocking failures.

Reforestation or 'Reductions to NSR' are areas successfully reforested
by natural regeneration or planting. Data for 1989/90 onwards are based
on surveyed and reported changes to NSR areas of public (Crown) land,
from ministry annual reports. For earlier years, estimates are based on
areas harvested and planted, and estimated rates of natural
regeneration, losses to fires, insects, diseases and restocking
failures.

## Silvicultural Treatments:
Detailed silviculture records are stored in the Reporting Silviculture
Updates and Land Status Tracking System (RESULTS;
<http://www.for.gov.bc.ca/his/results/>).

1.  Data for fertilization, pruning and spacing treatments was taken
    from Ministry of Forests Lands and Natural Resource Operations
    annual reports - Silviculture Accomplishments on Crown Land Funded
    by All Funding Sources by Forest Region (table 2.8) available here:
    <http://www.for.gov.bc.ca/hfp/silviculture/statistics/statistics.htm>.

2.  Forests for Tomorrow (FFT) planting information was taken from FFT
    annual accomplishment reporting published here:
    <http://lbis.forestpracticesbranch.com/LBIS/accomplishments>.

3.  Data for select seed was calculated using the following methodology:

    1.  Source data from 2 tables via DataBC: (1) RESULTS - Activity
        Treatment Units
        (<http://www.data.gov.bc.ca/dbc/catalogue/detail.page?config=dbc&P110=recorduid:174422>)
        and (2) RESULTS - Planting
        (<http://www.data.gov.bc.ca/dbc/catalogue/detail.page?config=dbc&P110=recorduid:173789>).
        These data are licensed under the BC Open Government License
        (<http://www.data.gov.bc.ca/local/dbc/docs/license/OGL-vbc2.0.pdf>);

    2.  Join Table 2 based on the field
        `activity_treatment_unit_id`(selecting only records where
        `ATU_COMPLETION_DATE` \>= 2006-04-01) and Table 1, retaining
        only matching records;

    3.  Using the joined file, exclude records where
        `ACTUAL_TREATMENT_AREA` = null and `ACTUAL_PLANTED_NUMBER` = 0;

    4.  Calculate pro-rated AREA for each planting:\
        ((NUMBER\_PLANTED/ACTUAL\_PLANTED\_NUMBER)\*ACTUAL\_TREATMENT\_AREA);

    5.  Calculate Fiscal Year based on `ATU_COMPLETION_DATE`;

    6.  Narrow to select seed only using categories in
        `GENETIC_CLASS_CODE` and
        `SUPERIOR_PROVIDENCE_IND`:(Genetic\_class = 'A') or
        (Genetic\_class = 'B' and SUPERIOR\_PROVIDENCE\_IND = 'Y');

    7.  Summarize pro-rated AREA by Fiscal Year.

An individual site may be treated more than once. Site preparation,
brushing and planting of B class seed is not included. Private
landowners are not required to report treatments. Fiscal years were
rounded down to calendar years in order to standardize data display
(e.g., 1980/1981 was considered to be 1980).

## Timber Volume Gains Expected from Silvicultural Treatments:
Timber volume gains (millions of cubic metres of wood) were estimated
using data submitted by licensees and the Ministry of Forests, Lands and
Natural resource operations to Reporting Silviculture Updates and Land
Status Tracking System (RESULTS;
<http://www.for.gov.bc.ca/his/results/>) and Seed Planning and Registry
System (SPAR; <http://www.for.gov.bc.ca/HTI/spar/>). Genetic gain data
are contained in the Ministry of Forests, Lands and Natural Resources
Operations Seed Planning and Registry System (SPAR;
<http://www.for.gov.bc.ca/HTI/spar/>).

Timber volume gains (cubic metres per hectare per year) were calculated
by multiplying the number of hectares treated, from Silvicultural
Treatments (see above), by the average annual growth rate multiplied by
65 (assumed rotation age) divided by 1,000,000. Site preparation,
brushing and planting of B class seed are not included as enhanced
silviculture treatments as they are considered to be the required
treatments post-harvest to achieve free growing conditions, these
activities are not considered to create incremental volume, and thus
gains from these treatments are considered to be baseline.

Annual growth rates per hectare:

-   0.50 m3 planting one to three years after logging (relative to
    satisfactory natural regeneration)

-   2.93 m3 backlog planting (relative to a NSR site, overgrown by
    competing vegetation)

-   0.40 m3 fertilizing

-   0.25 m3 spacing

-   0.41 m3 select seed (average gain in 2004/05, adjusted for lower and
    higher genetic quality in earlier and later years, respectively). In
    2009/2010 the select seed average gain was upped to 0.57m3 and in
    2010/2011 it was upped to 0.615 m3 based on the forest genetics
    council annual report actual and forecast genetic worth.

***

Published and Available On-Line at Environmental Reporting BC (July
2014):
<http://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>

Email correspondence to: envreportbc@gov.bc.ca
