---
topic: "land"
title: "Trends in Silviculture in B.C. (1987-2016)"
output: envreportbc::print_ver
---


```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
require(knitr)
library(readr) #read CSV file
library(dplyr) #data cleaning
library(tidyr) #wide to long df format
library(ggplot2) #plotting
library(envreportutils) #for soe_theme(), package from GitHub

opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=5)
read_chunk("silviculture.R")

# Set chart font to "" so uses pdf font
chart_font_web <- ""
```

```{r pre}

```

Managing and conserving the province's Crown forest and range resources
provides environmental, social and economic benefits to all British
Columbians. Silviculture---one of the primary means to enhance the social
and economic benefits from our forest resource---entails the manipulation
of forest and woodland vegetation in stands and on landscapes to meet
the diverse needs and values of landowners and society on a sustainable
basis. This indicator investigates the trends in four measures of
silviculture from 1987 to 2016.


-   **What silvicultural systems are used?** The three general
    categories of silvicultural systems used in B.C. are clearcutting,
    partial cutting and clearcutting with reserves. The area of Crown
    forest harvested annually averages 209,000 hectares each year.
    Clearcutting with reserves was the most common silvicultural system
    applied over the last 15 years.
-   **How much is reforested after disturbance?** The area of Crown land
    reforested annually is approximately the same as the area harvested
    two to five years earlier. Additional natural disturbance areas are
    reforested through government funded programs. For example, the
    [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    has funded 110,000 hectares of rehabilitation planting since 2005.
-   **What incremental silvicultural treatments have been done?**
    Investments in incremental silviculture to improve the growth and
    quality of future crop trees included fertilizing, pruning, spacing
    and using select seed (usually from seed orchards) for planting.
    Since 1987, silvicultural treatments totalled approximately 2.7
    million hectares.
-   **What volume gains will incremental silvicultural treatments
    yield?** Cumulative volume gains 65 years after making investments
    in incremental silviculture since 1987 are estimated at 92 million
    cubic metres (m^3^).

\vspace{30pt}

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
  
   \begingroup\centering{\textbf{What is Silviculture?}}
   
     \endgroup
     
    \vspace{1mm}
    
    \text{Silviculture is the art and science of controlling the establishment, growth,}     
    \text{composition, health and quality of forests and woodlands.}
 
 \vspace{3mm}    

    \text{Visit the \href{http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5}{Silviculture Program Page} to learn more about silviculture strategies } 
    
    \text{in British Columbia.}

  \end{minipage}
}
\end{center}

    
\newpage

## What Silvicultural Systems Are Used?

```{r silviculture, fig.height = 6, fig.width = 11}
```

-   A silvicultural system is a planned program of activities that
    encompasses how a stand of trees is harvested, regenerated, and
    managed over time. Selection of a silvicultural system depends on
    the ecological traits of the tree species, and on balancing the
    objectives of the landowner.
-   A variety of silvicultural systems are used in B.C. and can be
    grouped into three general categories: clearcutting, partial
    cutting, and clearcutting with reserves. Each name reflects the type
    of forest structure remaining after the initial harvest.
-   Since 1987, the area of Crown forest harvested annually has ranged
    from 150,000 hectares to 250,000 hectares, with an average of
    approximately 209,000 hectares each year.
-   From 1987-1996, a clearcutting silvicultural system was applied on
    88% of the area harvested. By 2012, clearcutting with reserves
    accounted for 85% of the area harvested on public forest land. While
    both systems remove the majority of the trees, clearcutting with
    reserves saves some trees within or outside the cutting boundary for
    other purposes such as wildlife habitat, water quality and visual
    landscapes.
    
\newpage

## How Much is Reforested After Disturbance?

```{r dist-refor, fig.height = 6, fig.width = 11}
```

-   Forests disturbed by timber harvests and other natural causes
    reforest naturally over time. Silviculture investments accelerate
    reforestation, increase timber supplies and restore ecological
    services sooner.
-   In 1987, explicit reforestation obligations on public land were
    introduced requiring holders of harvesting rights to reforest the
    areas they harvest. This led to planting a greater proportion of
    harvested areas.
-   In the early 1990’s, increased investments in site preparation,
    research that led to better planting methods, planting,
    rehabilitation planting and brushing ensured prompt restocking and
    the growth of desired trees.
-   Recent catastrophic wildfires and the mountain pine beetle epidemic
    have added large areas with below acceptable stocking. Since 2005,
    the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    has funded the reforestation of mountain pine beetle and wildfire
    impacted land not currently under legal reforestation obligations.
-   Forests disturbed by timber harvest can take 7 years or more to
    regenerate, an average of 1.9 years when planted and an average of
    5.5 years when left to reforest naturally. This is referred to as
    regeneration delay and accounts for a large amount of the gap
    between disturbance and reforestation area in the last 7 years.
    
\newpage

## What Incremental Silvicultural Treatments Have Been Done?

```{r treat, fig.height = 6, fig.width = 11}
```


-   Incremental silviculture is an investment in future timber
    production and environmental benefits from forests. Incremental
    silviculture only includes treatments that are not part of basic
    silviculture, where natural unimproved seed sources are used.
    Incremental silviculture can increase timber quantity and quality,
    manage forest health and fire risks, and improve specific habitats,
    water quality and visual landscapes. Incremental silviculture also
    creates employment opportunities for communities.
-   Since 1987, investments in incremental silviculture to improve the
    growth and quality of future crop trees included fertilizing
    (302,000 hectares), pruning (47,000 hectares), spacing (403,000
    hectares) and using select seed---usually from seed orchards---for
    planting (1,856,000 hectares). Over this period, silvicultural
    treatments totalled approximately 2.7 million hectares.
-   The use of select seed increases the rate of tree growth, increasing
    future timber volume, reducing constraints on harvesting adjacent
    areas and reducing the need for costly brushing treatments.
-   Fertilization has seen significant increases since the inception of
    the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    in 2005.
-   The [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    was established to respond to catastrophic wildfires and the
    mountain pine beetle epidemic. Since 2005, the program has funded
    planting on approximately 110,000 hectares of mountain pine beetle
    and wildfire impacted land not currently under legal reforestation
    obligations.
    
    

\newpage

## What Volume Gains Will Incremental Silvicultural Treatments Yield?

```{r gains, fig.height = 6, fig.width = 11}
```

-   Timber volume gains from incremental silviculture are the estimated
    volume gains---based on yield modeling predictions---in 65 years when
    compared to basic silviculture using natural unimproved seed
    sources.
-   Gains from incremental silviculture include increased short- and
    mid-term timber supply through spacing and fertilization,
    accelerated development of mature or old growth forest
    characteristics where needed for wildlife or biodiversity, higher
    wood quality through pruning, more pleasing visual landscapes and
    planting to make up the long-term timber supply.
-   In general, compared with natural regeneration, planting increases
    harvestable volume by about 15% without the use of select seed and
    by about 31% with the use of select seed.
-   Cumulative volume gains 65 years after making investments in
    incremental silviculture since 1987 are estimated at 92 million
    cubic metres (m^3^), based on fertilization (4.2 million m^3^),
    Spacing (6.5 million m^3^), select seed (60.1 million m^3^) and
    reforestation through the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    (20.9 million m^3^).

\newpage

## References and other useful links
-   [Ministry of Forests, Lands and Natural Resource Operations Annual
    Reports](http://www2.gov.bc.ca/gov/content?id=C1A06DF943D84858AFA8E755910E9D41)
-   [Reporting Silviculture Updates and Land Status Tracking System
    (RESULTS)](http://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC)
-   [Silviculture Program
    Page](http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5)
-   [Seed Planning and Registry Application
    (SPAR)](http://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)
-   [Forests for Tomorrow
    (FFT)](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)


## Data
\*By accessing these datasets, you agree to the license associated with
each file, as indicated below.

-   [Indicator data: Silviculture
    Systems](https://catalogue.data.gov.bc.ca/dataset/d3f00308-9d4d-4dc6-bdec-996124b61974) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Disturbances and
    Reforestation](https://catalogue.data.gov.bc.ca/dataset/fbc78ed7-1fc3-49a2-b90c-fb4e53884c41) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Silvicultural
    Treatments](https://catalogue.data.gov.bc.ca/dataset/fd81a75c-4059-49dc-8d54-90eeae0bf5c9) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Timber Volume
    Gains](https://catalogue.data.gov.bc.ca/dataset/6db9467f-22e8-48e7-b6ce-b3c3f169d949) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (December
2017):
<http://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>

Email correspondence to: envreportbc@gov.bc.ca

\newpage

# Glossary

- **Silviculture** The art and science of controlling the
    establishment, growth, composition, health and quality of forests
    and woodlands. Silviculture entails the manipulation of forest and
    woodland vegetation in stands and on landscapes to meet the diverse
    needs and values of landowners and society on a sustainable basis.

- **Clearcutting** The process of removing all trees, large and small,
    in a stand in one cutting operation. Clear-cutting with reserves is
    the removal of the majority of the trees saving some within or
    outside the cut boundary for other purposes such as wildlife
    habitat.

- **Seed tree** Trees left after reproduction cutting to provide seeds
    for natural regeneration. Trees are chosen for desired species,
    being healthy and and having good growth form.

- **Shelterwood** Harvesting trees in two or more cuttings that
    creates conditions where a new generation of seedlings can be
    established under the remaining timber without planting. This system
    is implemented in forests with species that can establish themselves
    in partial shade under the remaining mature trees.

- **Selection** Harvesting trees in a way that moves a forest stand
    towards an uneven-aged condition.

- **Clearcutting with reserves** The removal of the majority of the
    trees saving some within or outside the cut boundary for other
    purposes such as wildlife habitat.

- **Variable retention** Retaining forest structural elements for at
    least one rotation in order to preserve environmental values.

- **Ecosystem services** Benefits to humankind and wildlife such as
    the production of food, provisioning of clean water, decomposition
    of wastes, control of micro-climate, regulating nutrient cycles,
    pollination, spiritual and recreational sites.

- **Not satisfactorily restocked (NSR)** Productive forest land that
    has been denuded and has failed, partially or completely, to
    regenerate either naturally or by planting or seeding to the
    specified or desired free growing standards for the site.

- **Site preparation** Any action, related to reforestation, to create
    an environment favourable for survival of suitable trees during the
    first growing season. It may alter the ground cover, soil or
    microsite conditions, using biological, mechanical, or manual
    clearing, prescribed burns, herbicides, or a combination of methods.
    Both natural regeneration and planting may be improved through site
    preparation.

- **Planting** Establishing a forest by setting out seedlings,
    transplants or cuttings in an area.

- **Back-log planting** Planting that is overdue. In general, planting
    is considered backlog planting if more than 7 years have elapsed
    since a site was cleared (by harvesting, fire, insects or disease)
    in the Interior, and more than 3 years have elapsed on the Coast of
    British Columbia.

- **Brushing** A silvicultural activity done by chemical, manual,
    grazing, or mechanical means to control competing forest vegetation
    and reduce competition for space, light, moisture, and nutrients
    with crop trees or seedlings.

- **Incremental Silviculture** Application of cultural measures which will allow an increase in the 
    value or volume of the cut over and above the establishment of a fully 
    stocked stand after harvest.

- **Fertilizing** The addition of fertilizer to promote tree growth on
    sites deficient in one or more soil nutrients. Also used to improve
    the vigor of crop trees following juvenile spacing or commercial
    thinning.
    
- **Forests for Tomorrow** A silviculture program, established by the B.C. Provincial Government 
    in March 2005 to respond to catastrophic wildfires and the mountain pine 
    beetle epidemic.

- **Pruning** The manual removal, close to or flush with the stem, of
    side branches, live or dead, and of multiple leaders from standing,
    generally plantation-grown trees. Pruning is carried out to improve
    the market value of the final wood product by producing knot-free
    wood for the improvement of the tree or its timber.

- **Spacing** The removal of undesirable trees within a young stand to
    control stocking, to maintain or improve growth, to increase wood
    quality and value, or to achieve other resource management
    objectives.

- **Select seed** Seed that exhibits a higher level of one or more
    desired genetic traits (such as growth rate, form, wood density, and
    resistance to insects and disease) than wild seed collected from an
    average natural stand. This includes seed from tested parents
    growing in seed orchards and seed collected from natural stand
    superior provenances. Vegetative material for propagation, from
    production facilities using tested parents and from superior
    provenances, is included in the term select seed.

- **Natural regeneration** A change in forest structure and
    composition caused by fire, insects, wind, landslides and other
    natural processes.

- **Mid-term timber supply** The mid-term timber supply is a term that
    refers to that portion of the timber inventory that would be
    available for harvest within the middle of the normal management
    cycle.


\newpage

# METHODS

The [**R** code](http://www.r-project.org) for creating the graphs presented on this page is available on [GitHub](https://github.com/bcgov/EnvReportBC).

## Source Data

