---
topic: "land"
title: "Trends in Silviculture in B.C. (1987-2023)"
output: envreportutils.internal::print_ver
header-includes:
  - \usepackage{xcolor,colortbl}
  - \usepackage{longtable, booktabs}
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->
```

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
require(knitr)
library(readr) #read CSV file
library(dplyr) #data cleaning
library(tidyr) #wide to long df format
library(ggplot2) #plotting
library(envreportutils) #for soe_theme(), package from GitHub
library(here)

opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=5)
read_chunk("silviculture.R")

# Set chart font to "" so uses pdf font
chart_font_web <- ""
```

```{r pre}

```

Managing and conserving Crown forest resources provides environmental, social, and economic benefits to all British Columbians. Silviculture is a forest management tool used to control the establishment, growth, composition, and health of forests at different scales. Silviculture is used to enhance the social and economic benefits from forests to meet the diverse needs and values of B.C. citizens in a sustainable way. This indicator investigates the trends in four silviculture measures from 1987 to 2023.

-   **What silvicultural systems are used?** The three general
    categories of silvicultural systems used in B.C. are clearcutting,
    partial cutting, and clearcutting with reserves. Over the last 15 years, the area of Crown forest harvested annually averages 180,000 hectares. Clearcutting with reserves was the most common silvicultural system applied over this period.

-   **How much is reforested after disturbance?** The area of Crown land
    reforested annually is approximately the same as the area harvested
    two to five years earlier. Additional natural disturbance areas are reforested through government funded programs. Collectively, the Forest Investment Program (FIP) and its precursors (Forests for Tomorrow and Forest Carbon Initiative) have funded 282,000 hectares of rehabilitation planting since 2005.

-   **What incremental silvicultural treatments have been done?**
    Investments in incremental silviculture to improve the growth and
    quality of future crop trees included fertilizing, pruning, spacing
    and using select seed (usually from seed orchards) for planting.
    Since 1987, incremental silvicultural treatments totaled
    approximately 4 million hectares.


\vspace{30pt}

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
  
   \begingroup\centering{\textbf{What is Silviculture?}}
   
   \endgroup
     
    \vspace{1mm}
    
    \text{Silviculture is the art and science of controlling the establishment, }     
    \text{growth, composition, health and quality of forest vegetation.}
 
 \vspace{3mm}    

    \text{Visit the \href{http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5}{Silviculture Program Page} to learn more about silviculture } 
    
    \text{strategies in British Columbia.}

  \end{minipage}
}
\end{center}

\newpage

## What Silvicultural Systems Are Used?

-   A silvicultural system is a planned program of treatments during the whole life of a stand designed to achieve specific stand structural objectives. The three general categories of silvicultural systems used in B.C. are clearcutting, partial cutting, and clearcutting with reserves. Selection of a silvicultural system depends on the forest stand and site characteristics (e.g., tree species, forest health, terrain), and resource management objectives (e.g., timber, visual landscapes, wildlife habitat, social).

-   From 1987-1996, a clearcutting silvicultural system was applied on
    88% of the area harvested. By 2012, clearcutting with reserves
    accounted for 85% of the area harvested on public forest land. While
    both systems remove the majority of the trees, clearcutting with
    reserves saves some trees within or outside the cutting boundary for
    the purpose of managing for other values such as wildlife habitat,
    water quality or visual landscapes.
    
- Since 1987, the area of Crown forest harvested annually has ranged
    from 125,000 hectares to 256,000 hectares, with an average of
    approximately 207,000 hectares.


```{r silviculture, fig.height = 6, fig.width = 11}
```


\newpage

## How Much is Reforested After Disturbance?

- Forests disturbed by timber harvests and other natural causes
    reforest naturally over time. Silviculture investments accelerate
    reforestation, increase timber supplies and restore ecological
    services sooner.
    
-   Forests disturbed by timber harvest can take 7 years or more to
    regenerate. The average is 1.9 years when planted and 5.5 years when
    left to reforest naturally. This is referred to as regeneration
    delay and accounts for a large amount of the gap between disturbance
    and reforestation area in the last 7 years.

-   In 1987, requirements were introduced for rights holders to reforest public lands they harvest. This led to planting a greater proportion of harvested areas. In the early 1990's, increased investment in planning and research improved the rate and success of replanting efforts.

-   Recent catastrophic wildfires and the mountain pine beetle epidemic
    have added large areas with below acceptable stocking. Since 2005, the province has funded reforestation and rehabilitation of mountain pine beetle and wildfire impacted land not currently under legal reforestation obligations.


```{r dist-refor, fig.height = 6, fig.width = 11}
```


\newpage

## What Incremental Silvicultural Treatments Have Been Done?

-   Incremental silviculture is an investment in future timber
    production and environmental benefits from forests. Incremental
    silviculture only includes treatments that are not part of basic
    silviculture, where natural unimproved seed sources are used.
    Incremental silviculture can increase timber quantity and
    landscapes, manage forest health and fire risks, and improve
    specific habitats, water quality and visual landscapes. Incremental
    silviculture also creates employment opportunities for communities.

-   Since 1987, investments in incremental silviculture in B.C. included using select seed for planting (2,747,000 hectares), aerial fertilizing (508,000 hectares), spacing (427,000 hectares), and pruning (53,000 hectares).  Over this period, incremental silvicultural treatments totaled approximately 4 million hectares.

-   The use of select seed increases the rate of tree growth, increasing
    future timber volume, reducing constraints on harvesting adjacent
    areas and reducing the need for costly brushing treatments.

-   In 2021, the province committed to a consolidated [Forest Investment Program (FIP)](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/forest-investment-program") that combines incremental government-led forest investments previously managed by the Forest Carbon Initiative and the Forests for Tomorrow program. Under FIP, the province and its partners currently invest around $95 million a year to keep B.C. forests healthy and resilient. Every year since its inception, more than 40 million trees have been planted and over 30,000 hectares of forest have received a forest nutrient application.

```{r treat, fig.height = 6, fig.width = 11}
```

\newpage



## References and other useful links

-   [Reporting Silviculture Updates and Land Status Tracking System
    (RESULTS)](http://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC)
-   [Silviculture Program
    Page](http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5)
-   [Seed Planning and Registry Application
    (SPAR)](http://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)
-   [Forest Investment Program (FIP)](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/forest-investment-program)
-   [Forest Carbon
    Initiative](https://www2.gov.bc.ca/gov/content?id=3919ABD1BBC747069352CE0C5F84A37F)

## Data

\*By accessing these datasets, you agree to the license associated with
each file, as indicated below.

-   [Indicator Summary Data: Change in Silviculture Systems
    (1987-2023)](https://catalogue.data.gov.bc.ca/dataset/b3369823-d130-4e8d-bcca-55c8749fab40)
    (Licence: [Open Government Licence - British
    Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator Summary Data: Change in Disturbances and Reforestation
    (1987-2023)](https://catalogue.data.gov.bc.ca/dataset/b3369823-d130-4e8d-bcca-55c8749fab40)
    (Licence: [Open Government Licence - British
    Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator Summary Data: Change in Silvicultural Treatments
    (1987-2023)](https://catalogue.data.gov.bc.ca/dataset/b3369823-d130-4e8d-bcca-55c8749fab40)
    (Licence: [Open Government Licence - British
    Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))


For more information on this indicator or on Silviculture in British
Columbia contact the Resource Practices Branch at
[Forests.ForestPracticesBranchOffice\@gov.bc.ca](Forests.ForestPracticesBranchOffice@gov.bc.ca).

Published and Available On-Line at Environmental Reporting BC (May
2023): <http://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>

Email correspondence to:
[envreportbc\@gov.bc.ca](mailto:envreportbc@gov.bc.ca){.email}

\newpage

# Glossary

**Basic silviculture** Harvesting methods and silviculture operations
including seed collecting, site preparation, artificial and natural
regeneration, fertilization at the time of planting, brushing, spacing
and stand tending and other operations as are prescribed to be required
for the purpose of establishing a free growing crop of trees of a
commercially valuable species.

**Below acceptable stocking** Productive forest land that has been
denuded and has failed, partially or completely, to regenerate either
naturally or by planting or seeding to the specified or desired free
growing standards for the site.

**Brushing** A silvicultural activity done by chemical, manual, grazing,
or mechanical means to control competing forest vegetation and reduce
competition for space, light, moisture, and nutrients with crop trees or
seedlings.

**Clearcutting** The process of removing all trees, large and small, in
a stand in one cutting operation.

**Clearcutting with reserves** The removal of the majority of the trees
saving some within or outside the cut boundary for other purposes such
as wildlife habitat, water quality and visual landscapes.

**Fertilizing** The addition of fertilizer to promote tree growth on
sites deficient in one or more soil nutrients. Also used to improve the
vigor of crop trees following juvenile spacing or commercial thinning.

**Forests for Tomorrow** A silviculture program, established by the B.C.
Provincial Government in March 2005 to respond to catastrophic wildfires
and the mountain pine beetle epidemic.

**Incremental Silviculture** Silviculture other than basic silviculture.

**Mid-term timber supply** A term that refers to that portion of the
timber inventory that would be available for harvest within the middle
of the normal management cycle.

**Natural regeneration** A change in forest structure and composition
caused by fire, insects, wind, landslides and other natural processes.

**Partial cutting** A variety of silvicultural system in which a stand
may be cut to ensure regeneration. In a partial cutting system, only
some of the trees are felled during the harvesting phase. The selection
method may specify ‘removal’ or ‘leave’ trees. Some examples of
selection criteria are diameter, species, volume, age, height, disease,
or other damage.

**Planting** Establishing a forest by setting out seedlings, transplants
or cuttings in an area.

**Pruning** The manual removal, close to or flush with the stem, of side
branches, live or dead, and of multiple leaders from standing, generally
plantation-grown trees. Pruning is carried out to improve the market
value of the final wood product by producing knot-free wood for the
improvement of the tree or its timber.

**Rehabilitation planting** Where planting occurs on areas disturbed by
a natural event such as fire or insects which has not been salvaged in a
manner that would generate a legal reforestation obligation.

**Select seed** Seed collected from either orchards or natural stand
superior provenances. Select seed exhibits a higher level of improvement
in one or more desired genetic traits (such as growth, form, wood
density, and resistance to insects and disease) than wild seed collected
from an average natural stand.

**Site preparation** Any action, related to reforestation, to create an
environment favourable for survival of suitable trees during the first
growing season. It may alter the ground cover, soil or microsite
conditions, using biological, mechanical, or manual clearing, prescribed
burns, herbicides, or a combination of methods. Both natural
regeneration and planting may be improved through site preparation.

**Spacing** The removal of undesirable trees within a young stand to
control stocking, to maintain or improve growth, to increase wood
quality and value, or to achieve other resource management objectives.

\newpage

```{r child=here("silviculture", "methods-child.Rmd")}
```

