---
topic: "land"
title: "Trends in Silviculture in B.C. (1987-2016)"
output: envreportutils.internal::print_ver
---


```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}
require(knitr)
library(readr) #read CSV file
library(dplyr) #data cleaning
library(tidyr) #wide to long df format
library(ggplot2) #plotting
library(envreportutils) #for soe_theme(), package from GitHub

opts_chunk$set(echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE, error=FALSE, fig.height=5)
read_chunk("silviculture.R")

# Set chart font to "" so uses pdf font
chart_font_web <- ""
```

```{r pre}

```

Managing and conserving the province's Crown forest and range resources
provides environmental, social and economic benefits to all British
Columbians. Silviculture---one of the primary means to enhance the social
and economic benefits from our forest resource---entails the manipulation
of forest and woodland vegetation in stands and on landscapes to meet
the diverse needs and values of landowners and society on a sustainable
basis. This indicator investigates the trends in four measures of
silviculture from 1987 to 2016.


-   **What silvicultural systems are used?** The three general
    categories of silvicultural systems used in B.C. are clearcutting,
    partial cutting and clearcutting with reserves. The area of Crown
    forest harvested annually averages 209,000 hectares each year.
    Clearcutting with reserves was the most common silvicultural system
    applied over the last 15 years.
-   **How much is reforested after disturbance?** The area of Crown land
    reforested annually is approximately the same as the area harvested
    two to five years earlier. Additional natural disturbance areas are
    reforested through government funded programs. For example, the
    [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    has funded 110,000 hectares of rehabilitation planting since 2005.
-   **What incremental silvicultural treatments have been done?**
    Investments in incremental silviculture to improve the growth and
    quality of future crop trees included fertilizing, pruning, spacing
    and using select seed (usually from seed orchards) for planting.
    Since 1987, silvicultural treatments totalled approximately 2.7
    million hectares.
-   **What volume gains will incremental silvicultural treatments
    yield?** Cumulative volume gains 65 years after making investments
    in incremental silviculture since 1987 are estimated at 92 million
    cubic metres (m^3^).

\vspace{30pt}

\begin{center}
\fbox{
  \begin{minipage}[c]{12cm}
  
   \begingroup\centering{\textbf{What is Silviculture?}}
   
     \endgroup
     
    \vspace{1mm}
    
    \text{Silviculture is the art and science of controlling the establishment, growth,}     
    \text{composition, health and quality of forests and woodlands.}
 
 \vspace{3mm}    

    \text{Visit the \href{http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5}{Silviculture Program Page} to learn more about silviculture strategies } 
    
    \text{in British Columbia.}

  \end{minipage}
}
\end{center}

    
\newpage

## What Silvicultural Systems Are Used?

```{r silviculture, fig.height = 6, fig.width = 11}
```

-   A silvicultural system is a planned program of activities that
    encompasses how a stand of trees is harvested, regenerated, and
    managed over time. Selection of a silvicultural system depends on
    the ecological traits of the tree species, and on balancing the
    objectives of the landowner.
-   A variety of silvicultural systems are used in B.C. and can be
    grouped into three general categories: clearcutting, partial
    cutting, and clearcutting with reserves. Each name reflects the type
    of forest structure remaining after the initial harvest.
-   Since 1987, the area of Crown forest harvested annually has ranged
    from 150,000 hectares to 250,000 hectares, with an average of
    approximately 209,000 hectares each year.
-   From 1987-1996, a clearcutting silvicultural system was applied on
    88% of the area harvested. By 2012, clearcutting with reserves
    accounted for 85% of the area harvested on public forest land. While
    both systems remove the majority of the trees, clearcutting with
    reserves saves some trees within or outside the cutting boundary for
    other purposes such as wildlife habitat, water quality and visual
    landscapes.
    
\newpage

## How Much is Reforested After Disturbance?

```{r dist-refor, fig.height = 6, fig.width = 11}
```

-   Forests disturbed by timber harvests and other natural causes
    reforest naturally over time. Silviculture investments accelerate
    reforestation, increase timber supplies and restore ecological
    services sooner.
-   In 1987, explicit reforestation obligations on public land were
    introduced requiring holders of harvesting rights to reforest the
    areas they harvest. This led to planting a greater proportion of
    harvested areas.
-   In the early 1990’s, increased investments in site preparation,
    research that led to better planting methods, planting,
    rehabilitation planting and brushing ensured prompt restocking and
    the growth of desired trees.
-   Recent catastrophic wildfires and the mountain pine beetle epidemic
    have added large areas with below acceptable stocking. Since 2005,
    the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    has funded the reforestation of mountain pine beetle and wildfire
    impacted land not currently under legal reforestation obligations.
-   Forests disturbed by timber harvest can take 7 years or more to
    regenerate, an average of 1.9 years when planted and an average of
    5.5 years when left to reforest naturally. This is referred to as
    regeneration delay and accounts for a large amount of the gap
    between disturbance and reforestation area in the last 7 years.
    
\newpage

## What Incremental Silvicultural Treatments Have Been Done?

```{r treat, fig.height = 6, fig.width = 11}
```


-   Incremental silviculture is an investment in future timber
    production and environmental benefits from forests. Incremental
    silviculture only includes treatments that are not part of basic
    silviculture, where natural unimproved seed sources are used.
    Incremental silviculture can increase timber quantity and quality,
    manage forest health and fire risks, and improve specific habitats,
    water quality and visual landscapes. Incremental silviculture also
    creates employment opportunities for communities.
-   Since 1987, investments in incremental silviculture to improve the
    growth and quality of future crop trees included fertilizing
    (302,000 hectares), pruning (47,000 hectares), spacing (403,000
    hectares) and using select seed---usually from seed orchards---for
    planting (1,856,000 hectares). Over this period, silvicultural
    treatments totalled approximately 2.7 million hectares.
-   The use of select seed increases the rate of tree growth, increasing
    future timber volume, reducing constraints on harvesting adjacent
    areas and reducing the need for costly brushing treatments.
-   Fertilization has seen significant increases since the inception of
    the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    in 2005.
-   The [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    was established to respond to catastrophic wildfires and the
    mountain pine beetle epidemic. Since 2005, the program has funded
    planting on approximately 110,000 hectares of mountain pine beetle
    and wildfire impacted land not currently under legal reforestation
    obligations.
    
    

\newpage

## What Volume Gains Will Incremental Silvicultural Treatments Yield?

```{r gains, fig.height = 6, fig.width = 11}
```

-   Timber volume gains from incremental silviculture are the estimated
    volume gains---based on yield modeling predictions---in 65 years when
    compared to basic silviculture using natural unimproved seed
    sources.
-   Gains from incremental silviculture include increased short- and
    mid-term timber supply through spacing and fertilization,
    accelerated development of mature or old growth forest
    characteristics where needed for wildlife or biodiversity, higher
    wood quality through pruning, more pleasing visual landscapes and
    planting to make up the long-term timber supply.
-   In general, compared with natural regeneration, planting increases
    harvestable volume by about 15% without the use of select seed and
    by about 31% with the use of select seed.
-   Cumulative volume gains 65 years after making investments in
    incremental silviculture since 1987 are estimated at 92 million
    cubic metres (m^3^), based on fertilization (4.2 million m^3^),
    Spacing (6.5 million m^3^), select seed (60.1 million m^3^) and
    reforestation through the [Forests for Tomorrow
    program](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)
    (20.9 million m^3^).

\newpage

## References and other useful links
-   [Ministry of Forests, Lands and Natural Resource Operations Annual
    Reports](http://www2.gov.bc.ca/gov/content?id=C1A06DF943D84858AFA8E755910E9D41)
-   [Reporting Silviculture Updates and Land Status Tracking System
    (RESULTS)](http://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC)
-   [Silviculture Program
    Page](http://www2.gov.bc.ca/gov/content?id=C9B072DFBCA445FF9AE0CE64ADA5FAA5)
-   [Seed Planning and Registry Application
    (SPAR)](http://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)
-   [Forests for Tomorrow
    (FFT)](http://www2.gov.bc.ca/gov/content?id=EDA5DA85D4EA43BEBC5868ACA132BF2C)


## Data
\*By accessing these datasets, you agree to the license associated with
each file, as indicated below.

-   [Indicator data: Silviculture
    Systems](https://catalogue.data.gov.bc.ca/dataset/d3f00308-9d4d-4dc6-bdec-996124b61974) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Disturbances and
    Reforestation](https://catalogue.data.gov.bc.ca/dataset/fbc78ed7-1fc3-49a2-b90c-fb4e53884c41) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Silvicultural
    Treatments](https://catalogue.data.gov.bc.ca/dataset/fd81a75c-4059-49dc-8d54-90eeae0bf5c9) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator data: Timber Volume
    Gains](https://catalogue.data.gov.bc.ca/dataset/6db9467f-22e8-48e7-b6ce-b3c3f169d949) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (December
2017):
<http://www.env.gov.bc.ca/soe/indicators/land/silviculture.html>

Email correspondence to: envreportbc@gov.bc.ca

\newpage

# Glossary

**Basic silviculture** Harvesting methods and silviculture
operations including seed collecting, site preparation, artificial and
natural regeneration, brushing, spacing and stand tending and other
operations as are prescribed to be required for the purpose of
establishing a free growing crop of trees of a commercially valuable
species.

**Below acceptable stocking** Productive forest land that has been
denuded and has failed, partially or completely, to regenerate either
naturally or by planting or seeding to the specified or desired free
growing standards for the site.

**Brushing** A silvicultural activity done by chemical, manual, grazing,
or mechanical means to control competing forest vegetation and reduce
competition for space, light, moisture, and nutrients with crop trees or
seedlings.

**Clearcutting** The process of removing all trees, large and small, in
a stand in one cutting operation.

**Clearcutting with reserves** The removal of the majority of the trees
saving some within or outside the cut boundary for other purposes such
as wildlife habitat, water quality and visual landscapes.

**Fertilizing** The addition of fertilizer to promote tree growth on
sites deficient in one or more soil nutrients. Also used to improve the
vigor of crop trees following juvenile spacing or commercial thinning.

**Forests for Tomorrow** A silviculture program, established by the B.C.
Provincial Government in March 2005 to respond to catastrophic wildfires
and the mountain pine beetle epidemic.

**Incremental Silviculture** Silviculture other than basic silviculture.

**Mid-term timber supply** A term that refers to that portion of the timber inventory that would be available
for harvest within the middle of the normal management cycle.

**Natural regeneration** A change in forest structure and composition
caused by fire, insects, wind, landslides and other natural processes.

**Partial cutting** A variety of silvicultural system in which a stand
may be cut to ensure regeneration. In a partial cutting system, only
some of the trees are felled during the harvesting phase. The selection
method may specify ‘removal’ or ‘leave’ trees. Some examples of
selection criteria are diameter, species, volume, age, height, disease,
or other damage.

**Planting** Establishing a forest by setting out seedlings, transplants
or cuttings in an area.

**Pruning** The manual removal, close to or flush with the stem, of side
branches, live or dead, and of multiple leaders from standing, generally
plantation-grown trees. Pruning is carried out to improve the market
value of the final wood product by producing knot-free wood for the
improvement of the tree or its timber.

**Rehabilitation planting** Where planting occurs on areas disturbed by
a natural event such as fire or insects which has not been salvaged in a
manner that would generate a legal reforestation obligation.

**Select seed** Seed collected from either orchards or natural stand
superior provenances. Select seed exhibits a higher level of improvement
in one or more desired genetic traits (such as growth, form, wood
density, and resistance to insects and disease) than wild seed collected
from an average natural stand.

**Site preparation** Any action, related to reforestation, to create an
environment favourable for survival of suitable trees during the first
growing season. It may alter the ground cover, soil or microsite
conditions, using biological, mechanical, or manual clearing, prescribed
burns, herbicides, or a combination of methods. Both natural
regeneration and planting may be improved through site preparation.

**Spacing** The removal of undesirable trees within a young stand to
control stocking, to maintain or improve growth, to increase wood
quality and value, or to achieve other resource management objectives.

\newpage

# METHODS

The [**R** code](http://www.r-project.org) for creating the graphs presented on this page is available on [GitHub](https://github.com/bcgov/EnvReportBC).

## Source Data

The Silviculture analysis uses data from:

1.  [Reporting Silviculture Updates and Land Status Tracking System (RESULTS)](https://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC>) and

2.  [Seed Planning and Registry System (SPAR)](https://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)


Wherever possible, the analysis used public data layers from the B.C.
Geographic Warehouse[^1] released under either the Open Government Licence---British Columbia or the B.C. Crown Copyright Licence:

-   **Opening\_VW**: [WHSE\_FOREST\_VEGATATION\_OPENING\_VW](https://catalogue.data.gov.bc.ca/dataset/f6d17c00-d4b9-401b-a20e-ef7e37b6d5e3) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   **Seedlot**: [WHSE\_FOREST\_VEGATATION\_SEED\_SEEDLOT](https://catalogue.data.gov.bc.ca/dataset/686e10d6-78e3-4779-9a8b-08015a5c070d) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   **Veglot**: [WHSE\_FOREST\_VEGETATION.SEED\_VEG\_LOT](https://catalogue.data.gov.bc.ca/dataset/2456f4ed-7e47-438f-9ab5-711268c92075) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   **Activity Treatment Unit**: [WHSE\_FOREST\_VEGATATION\_RSLT\_ACTIVITY\_TREATMENT\_UNIT](https://catalogue.data.gov.bc.ca/dataset/07cabbdf-d7bf-4c50-919d-5b7d80086ef5) (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

-   **Planting**: WHSE\_FOREST\_VEGATATION\_RSLT\_PLANTING\_RSLT

-   **Seedlot Genetic Worth**: SEEDLOT\_GENETIC\_WORTH

-   **Org Unit**: Appendix I – Lookup Tables

-   **CoastInteriorOrgUnit**: Appendix I – Lookup Tables

-   **SILV\_FUND\_SRC\_CODE**: Appendix I – Lookup Tables

The intent is to move the two data layers above that are currently not publicly available into the B.C. Geographic Warehouse so that the raw data and source code
can be published to allow interested parties to repeat the analysis.

## Fiscal Year

Fiscal year will be display as follows:

-   Formula used: Fiscal Yr:```IIf(Month([DATE])<4,Year([DATE])-1,Year([DATE]))```

-   Example: 1987-04-01 and 1988-03-31 Fiscal Year = 1988

## Baseline Make Table Queries

All queries are built in Microsoft Access 2007. Make Table queries are
used to create baseline data tables for increased efficiency and data
auditability. Note that all data reported after April 1, 1970 is added
to the baseline but the data are subset to more recent years in the
final queries.

## Silvicultural Systems

*1a-Make Opening (outcome: 1a-Opening)*

Extracts all the required attributes from the Opening\_VW to create
the 1a-Opening table to get a unique list of all openings.

*2a-Make Raw Silv System (outcome: 2a-Raw Silv System)*

Lists all completed disturbance reporting from the Activity Treatment
Unit table.

  **Field**          **Criteria**   **Details**
  ------------------ -------------- ----------------------------
  SILV\_BASE\_CODE   DN             Disturbance Reporting Only
  RESULTS\_IND       Y              Completed activities only

*2a-Detailed Silv System and Fiscal Date*

Joins the 2a-Raw Silv System with the 1a-Make Opening and creates
some new calculated attributes.

--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                                      **Criteria**                                                                                                                                                                        **Details**
  ---------------------------------------------- ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- ---------------------------------------------------------------------------------------
  NewSilv                                        ```NewSilv: ([DISTURBANCE_CODE] In ("L","S") And [SILV_SYSTEM_CODE] Is Null,"CLEAR", [SILV_SYSTEM_CODE])```                                                                         Identifies disturbance records with a silv system code that are Logging or Salvage and identifies them as a clearcut
  
  
  Dist Start Yr                                  ```IIf(Month([ATU_START_DATE])<4, Year([ATU_START_DATE])-1, Year([ATU_START_DATE]))```                                                                                              Changes the disturbance start date to calendar year
  
  
  Dist End Yr                                    ```IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))```                                                                               Changes the disturbance end date to calendar year
  
  
  Disturbance Type                               ```IIf([DISTURBANCE_CODE] In ("L","S") And [2a-Raw Silv System]![SILV_SYSTEM_CODE] Is Null Or [2a-Raw Silv System]![SILV_SYSTEM_CODE] Is Not Null,"Harvest","Not Harvest")```       Identifies disturbance records with a Harvest or Not Harvest identifier
  
--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*2b-ATU DN Detail Best Date*

Joins the 2a-Detailed Silv System and Fiscal Date NewSilv with the
following Silv System Code lookup table and defines the best fiscal year
to report the treatments to.

  **SILV\_SYSTEM\_CODE**   **Silviculture System**
  ------------------------ ----------------------------
  CCRES                    Clearcutting with reserves
  CLEAR                    Clearcutting
  COPPI                    Partial cutting
  IMCUT                    Partial cutting
  PATCT                    Partial cutting
  RETEN                    Clearcutting with reserves
  SEEDT                    Partial cutting
  SELEC                    Partial cutting
  SHELT                    Partial cutting


-------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                    **Criteria**                                                         **Details**
  ---------------------------- -------------------------------------------------------------------- --------------------------------------------------------------------------
  Best Dist End Yr             ```IIf([Dist End Yr] Is Null,[Dist Start Yr],[Dist End Yr])```       Populates missing disturbance ends years with the disturbance start year

-------------------------------------------------------------------------------------------------------------------------------------------------------------------------


*2c-Harvest Sum of DN Area*

Summarizes 2b-ATU DN Detail Best Date to calculate the disturbance
area in an opening.

  **Field**                **Criteria**             **Details**
  ------------------------ ------------------------ ----------------
  Disturbance Type         = "Harvest""             Group by (Row)
  OPENING\_ID                                       Group by (Row)
  LastOfBest Dist End Yr   LAST(Best Dist End Yr)   Group by (Row)
  OPENING\_GROSS\_AREA                              Group by (Row)
  DN\_Area                 TREATMENT\_AMOUNT        Sum

*2d-Harvest Best Area*

Calculates the best harvest area using 2c-Harvest Sum of DN Area.

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                               **Criteria**                                                                                       **Details**
  --------------------------------------- -------------------------------------------------------------------------------------------------- -----------------------------------------------------------------------------------------------------------------------------
  DN Area                                 <0                                                                                 
  
  Best Harv Area                          ```IIf([DN Area]<=[OPENING_GROSS_AREA], [DN Area],[OPENING_GROSS_AREA])```                          Where the sum of the DN Area is greater than the opening gross area, the opening gross area is used as the disturbance area

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- 

*2e-Harvest Silv System Area*

Summarizes 2b-ATU DN Detail Best Date to calculate the disturbance
area in an opening.

  **Field**                **Criteria**   **Details**
  ------------------------ -------------- -------------------
  Disturbance Type         = "Harvest""   Group by (Row)
  OPENING\_ID                             Group by (Row)
  LastOfBest Dist End Yr                  Group by (Row)
  SOF Silv System                         Group by (Column)
  TREATMENT\_AMOUNT        Sum            Sum

*2f-Make Silv System Ratio*

Joins the 2e-Harvest Silv System Area with the 2d-Harvest Best Area
and creates some new calculated attributes.

  **Field**      **Criteria**                         **Details**
  -------------- ------------------------------------ -------------
  CCRES\_pct     Round((\[CCRES\]/\[DN Area\]),2)     
  CLEAR\_pct     Round((\[CLEAR\]/\[DN Area\]),2)     
  PARTIAL\_pct   Round((\[PARTIAL\]/\[DN Area\]),2)   

*2g-Harvest Silv System Best Area*

Joins the 2f-Make Silv System Ratio with the 2d-Harvest Best Area
and creates some new calculated attributes.

  **Field**       **Criteria**                           **Details**
  --------------- -------------------------------------- -------------
  Final CCRES     \[CCRES\_pct\]\*\[Best Harv Area\]     
  Final CLEAR     \[CLEAR\_pct\]\*\[Best Harv Area\]     
  Final PARTIAL   \[PARTIAL\_pct\]\*\[Best Harv Area\]   

*2h-Final Silv System*

Summarizes 2g-Harvest Silv System Best Area to calculate the
disturbance area in an opening by silvicultural system.

  **Field**                **Criteria**   **Details**
  ------------------------ -------------- ----------------
  LastOfBest Harv End Yr   &gt;= 1987     Group by (Row)
  Final CCRES                             Sum
  Final CLEAR                             Sum
  Final PARTIAL                           Sum


## Disturbances and Reforestation

*2i-Disturbance Sum of DN Area*

Summarizes 2b-ATU DN Detail Best Date to calculate the natural
disturbance area in an opening.

  **Field**                **Criteria**             **Details**
  ------------------------ ------------------------ ----------------
  Disturbance Type         “Not Harvest”            Group by (Row)
  OPENING\_ID                                       Group by (Row)
  LastOfBest Dist End Yr   LAST(Best Dist End Yr)   Group by (Row)
  OPENING\_GROSS\_AREA                              Group by (Row)
  DN Area                  TREATMENT\_AMOUNT        Sum

*2j-Disturbance Best Area*

Calculates the best harvest date and area using the 1a-Opening and
2e-Summary DN Area.

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                                  **Criteria**                                                                             **Details**
  ------------------------------------------ ---------------------------------------------------------------------------------------- -----------------------------------------------------------------------------------------------------------------------------------------------------
  Best Disturb Area                          ```IIf([DN Area]<= [OPENING_GROSS_AREA], [DN Area], [OPENING_GROSS_AREA])```              Where the sum of the DN Area is less than the opening gross area, the DN Area is used as the Best Disturb area, else the opening gross area is used

----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*2k-Final Disturbance by Year*

Summarises 2j-Disturbance Best Area by Year.

  **Field**                **Criteria**       **Details**
  ------------------------ ------------------ ----------------
  LastOfBest Dist End Yr   Where &gt;= 1987   Group by (Row)
  Best Disturb Area                           Sum

*2l-All Denudation*

Joins the 2h-Final Silv System with the 2k-Final Disturbance by Year
to sum the total harvested and total natural disturbance areas by fiscal
year.

  **Field**           **Criteria**                                                                 **Details**
  ------------------- ---------------------------------------------------------------------------- -------------
  Total Harvested     ```[Final CCRES]+[Final CLEAR]+[Final PARTIAL]```                      
  Total Disturbance   ```[Final CCRES]+[Final CLEAR]+[Final PARTIAL]+[Not Harvest Area]```  

The reforestation number is calculated in the Trends in Genetic Resource
Conservation and Management of B.C. Forests indicator analysis. Details
on how the reforestation number was calculated are available in the
indicators methods document (query - 3g-Assess Regen Type – Set
Threshold).

## Silviculture Treatments

*6a-Make Fertilization (outcome: 6a-Fertilization)*

Joins the Activity Treatment Unit and Opening\_VW tables using the
Opening\_ID, joins the Activity Treatment Unit to the Org Unit using the
ORG\_UNIT\_NO and joins the Org Unit to the CoastInteriorOrgUnit using
the ORG\_UNIT\_CODE to get a complete list of fertilization activities.

  **Field**               **Criteria**    **Details**
  ----------------------- --------------- -------------
  SILV\_BASE\_CODE        FE              
  SILV\_TECHNIQUE\_CODE   Is Null or CA   
  TREATMENT\_AREA         &gt;0           

*6b-Fertilization with Coast Interior*

Joins 6a-Fertilization with 1a-Opening using the Opening\_ID, joins
6a-Fertilization with SILV\_FUND\_SRCE\_CODE using the
SILV\_FUND\_SRCE\_CODE, and joins the 1a-Opening with the Org Unit using
the ORG\_UNIT\_NO. These tables are joined to summarize fertilization
volume gains for the coast and interior (two different calculations) by
fiscal year.

  --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                                 **Criteria**                                                                                                        **Details**
  ----------------------------------------- ------------------------------------------------------------------------------------------------------------------- -------------------------------------------------
  Fert Yr                                   ```IIf(Month([ATU_COMPLETION_DATE]), 4,Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))```   

  Fertilization Volume Gain                 ```IIf([Coast_interior]="I", [TREATMENT_AMOUNT]*0.15*65, [TREATMENT_AMOUNT]*0.3*65)```                               Interior = 0.15m^3^/ha/yr Coast = 0.3m^3^/ha/yr
  
  ---------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*6c-Fertilization with Volume Gains – BC*

Summarizes 6b-Fertilization with Coast Interior Fertilization Area and
Fertilization Volume Gain by Fiscal Year.

  **Field**                   **Criteria**         **Details**
  --------------------------- -------------------- ----------------
  Fert Yr                     Between &gt;= 1970   Group by (Row)
  Fertilization Area          TREATMENT\_AMOUNT    Sum
  Fertilization Volume Gain                        Sum
  Category                    < > Legal            Where

*7a-Make Spacing (outcome: 7a-Spacing)*

Joins the Activity Treatment Unit and Opening\_VW tables using the
Opening\_ID to summarize spacing activities.

  **Field**            **Criteria**      **Details**
  -------------------- ----------------- -----------------------------
  SILV\_BASE\_CODE     JS                
  SILV\_METHOD\_CODE   < > LAYOT         Ignore preparatory activity
  TREATMENT\_AREA      &gt;0             

*7b-Spacing Volume Gains*

Joins 7a-Spacing with 1a-Opening using the Opening\_ID, joins 7a-Spacing
with SILV\_FUND\_SRCE\_CODE using the SILV\_FUND\_SRCE\_CODE, and joins
the 1a-Opening with the Org Unit using the ORG\_UNIT\_NO. These tables
are joined to summarize spacing volume by fiscal year.


--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**           **Criteria**                                                                                                          **Details**
  ------------------- --------------------------------------------------------------------------------------------------------------------- ---------------------------------------------------
  Incremental
  
  JS Gains            ```(0.25*[TREATMENT_AMOUNT])*65```                                                                                    Gains = .25m^3^/ha/yr
  
  JS Yr               ```JS Yr: IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))``` 
  
  Category            ```Not in (Legal, Other)```                                                                                           Excludes any legally obligated spacing activities
  
----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  
  
*7c-Spacing with Volume Gains – BC*

Summarizes 7b-Spacing with Volume Gains Spacing Area and Incremental JS
Gains by JS Year.

  **Field**              **Criteria**        **Details**
  ---------------------- ------------------- ----------------
  JS Yr                  &gt;= 1970          Group by (Row)
  Spacing Area           TREATMENT\_AMOUNT   Sum
  Incremental JS Gains                       Sum
  Category               &lt;&gt;“Legal”     Where

*8a-Make Pruning (outcome: 8a-Pruning)*

Joins the Activity Treatment Unit and Opening\_VW tables using the
Opening\_ID to summarize pruning activities.

  **Field**            **Criteria**      **Details**
  -------------------- ----------------- ------------------------------
  SILV\_BASE\_CODE     PR                
  SILV\_METHOD\_CODE   &lt; &gt; LAYOT   Removes preparatory activity
  TREATMENT\_AREA      &gt;0             

*8b-Pruning Area*

Joins 8a-Pruning with 1a-Opening using the Opening\_ID, joins 8a-Pruning
with SILV\_FUND\_SRCE\_CODE using the SILV\_FUND\_SRCE\_CODE, and joins
the 1a-Opening with the Org Unit using the ORG\_UNIT\_NO. These tables
are joined to summarize pruning volume by fiscal year.

--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**   **Criteria**                                                                                                               **Details**
  ----------- -------------------------------------------------------------------------------------------------------------------------- -------------
  PR Yr       ```PR Yr: IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE\]))```   
  
--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*8d-Pruning – BC*

Summarizes 8b-Pruning Area Pruning Area by PR Year.

  **Field**      **Criteria**        **Details**
  -------------- ------------------- ----------------
  PR Yr Yr       &gt;= 1970          Group by (Row)
  Pruning Area   TREATMENT\_AMOUNT   Sum
  Category       &lt;&gt;“Legal”     Where

*1a-Make Opening (outcome: 1a-Opening)*

Extracts the all the required attributes from the Opening\_VW to create
the 1a-Opening table to get a unique list of all openings.

*1b-Make Planting Activities (outcome: 1b-Planting Activities)*

Joins the Activity Treatment Unit and Opening\_VW tables using the
Opening\_ID to summarize planting activities.

  **Field**          **Criteria**   **Details**
  ------------------ -------------- ----------------------------
  SILV\_BASE\_CODE   PL             
  ACT\_PLANTED\_NO   &gt;0          Must contain planted trees

*1c-Make Planted Trees (outcome: 1b-Planted Trees)*

Joins the Activity Treatment Unit and Opening\_VW tables using the
Opening\_ID and joins the Activity Treatment Unit and Planting RSLT
using the ACTIVITY\_TREATMENT\_UNIT\_ID to summarize planted trees.

  **Field**          **Criteria**   **Details**
  ------------------ -------------- ----------------------------
  SILV\_BASE\_CODE   PL             
  ACT\_PLANTED\_NO   &gt;0          Must contain planted trees

*5a-Backlog Planting*

Joins the 1a-Opening to the 1b-Planted Activities using the Opening\_ID
and joins the 1b-Planted Activities to the SILV\_FUND\_SRCE\_CODE using
the SILV\_FUND\_SRCE\_CODE to summarize planting activities.

--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                **Criteria**                                                                                                        **Details**
  ------------------------ ------------------------------------------------------------------------------------------------------------------- --------------------------------------------------------
  Category                 ```FFT```                                                                                                            FFT is used to identify the post-2005 backlog planting
  
  Backlog Planted Yr       ```IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))```
  
  Backlog PL Volume Gain   ```(2.93*[TREATMENT_AMOUNT])*65```                                                                                   Gains = 2.93 m^3^/ha/yr
  
--------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*5b- Backlog Planting with Volume Gains*

Summarizes 8b-Pruning Area Pruning Area by PR Year.

  **Field**                **Criteria**               **Details**
  ------------------------ -------------------------- ----------------
  Backlog Planted Yr       &gt;= 1970                 Group by (Row)
  Backlog Planting Area    TREATMENT\_AMOUNT          Sum
  Backlog PL Volume Gain                              Sum
  Category                 Not In ("Legal","Other")   Where

*3a-Planting Details with Seedlot*

Joins the 1a-Opening to the 1b-Planted Activities using the opening\_ID,
joins the 1b-Planted Activities to the 1c-Planted Trees using the
ACTIIVTY\_TREATMENT\_UNIT\_ID and joins the 1c-Planted Trees to the
Seedlot using the SEEDLOT\_NUMBER to summarize planted tree details
needed to calculate genetic gains.


-----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                                       **Criteria**                                                                                                                                                                                                                                                                        **Details**
  ----------------------------------------------- ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- -------------------------------------------------
  Disturb Start Yr                                ```IIf(Month([DISTURBANCE_START_DATE]<4, Year([DISTURBANCE_START_DATE])-1, Year([DISTUBANCE_START_DATE]))```                                                                                                                                                                         Converts the Disturbance Start Date to Fiscal year
  
  Disturb End Yr                                  ```IIf(Month([DISTURBANCE_END_DATE]<4, Year([DISTURBANCE_END_DATE])-1, Year([DISTURBANCE_END_DATE]))```                                                                                                                                                                              Converts the Disturbance End Date to Fiscal year
  
  Planted Yr                                      ```IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))```                                                                                                                                                                                Converts the Planting Date to Fiscal year
  
  SeedClass                                       ```IIf([GENETIC_CLASS_CODE]="B" And [SUPERIOR_PRVNC_IND]="Y","B+", IIf([GENETIC_CLASS_CODE]="B" And [SUPERIOR_PRVNC_IND]="N","B", IIf([GENETIC_CLASS_CODE]="B" And [SUPERIOR_PRVNC_IND] Is Null,"B", IIf([GENETIC_CLASS_CODE]="A","A", IIf([GENETIC_CLASS_CODE] Is Null,"B")))))```  Assign Seed Class
  
  ProratePLArea                                   ```Round((([1c-Planted Trees]![NUMBER_PLANTED]/ [1c-Planted Trees]![ACT_PLANTED_NO])* [1c-Planted Trees]![TREATMENT_AMOUNT]),2)```                                                                                                                                                   Prorated species planted by planted area
  
  ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*4a-Calculate Genetic Worth G Gains by Opening*

Summarizes 3a-Planting Details with Seedlot by Opening\_ID where the
GENETIC\_WORTH\_CODE = “G” or the SeedClass = “A” OR “B+”.

  **Field**              **Criteria**   **Details**
  ---------------------- -------------- -----------------------------------------------------
  GENETIC\_WORTH\_CODE   G              Gain
  SeedClass              In (A, B+)     “A”=Orchard and “B+”=Planted Natural Stand Superior
  
*4b-Calculate Genetic Worth G Gains by Opening*

Calculates the Productivity Gain for each opening in the 4a-Calculate
Genetic Worth G Gains by Opening table.

  --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                           **Criteria**                                                                                                                                                                                     **Details**
  ----------------------------------- ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------ -----------------------------------------------------------
  Productivity Gain                   ```GProductivity Gain: IIf([PlantedYr]<=2009, 0.41*[ProratePLArea]\*65, IIf([PlantedYr]=2010, 0.53*[ProratePLArea]*65, IIf([PlantedYr]>2010, 0.615*[ProratePLArea]*65)))```                      < = 2009: 0.41m^3^/ha/yr  2010: 0.53m^3^/ha/yr >2010: 0.615m^3^/ha/yr
  
  --------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------                 


*4c-Summary Genetic Gains from Planting BC*

Takes 4b-Calculate Genetic Productivity G Gains by Opening and
summarizes gains by area and volume.

  **Field**            **Criteria**          **Details**
  -------------------- --------------------- ----------------
  Planted Yr           &gt;= 1970            Group by (Row)
  Select Seed Area     (ProratePLArea)       Sum
  Select Seed Volume   (Productivity Gain)   Sum

*9a-Silvicuture Treatments BC Area*

Joins the following tables to report all of the silviculture treatment
area summaries in one table:

-   5b-Backlog Planting with Volume Gains BC;

-   4c-Summary Genetic Gains from Planting BC;

-   6c-Fertilization with Volume Gains – BC;

-   7c-Spacing with Volume Gains – BC; and

-   8d-Pruning – BC.

*9b-Silvicuture Treatments BC Area*

Joins the following tables to report all of the silviculture treatment
area summaries in one table:

-   5b-Backlog Planting with Volume Gains BC;

-   4c-Summary Genetic Gains from Planting BC;

-   6c-Fertilization with Volume Gains – BC;

-   7c-Spacing with Volume Gains – BC; and

-   8d-Pruning – BC.

## Appendix I – Lookup tables


### Org Unit

  **ORG\_UNIT\_NO**   **ORG\_UNIT\_CODE**   **ROLLUP\_REGION\_CODE**
  ------------------- --------------------- --------------------------
  1826                DCC                   RCB
  15                  DCK                   RSC
  43                  DCR                   RWC
  1828                DCS                   RTO
  42                  DDC                   RNO
  46                  DFN                   RNO
  1827                DHW                   RSI
  34                  DJA                   ROM
  21                  DKA                   RTO
  32                  DKM                   RSK
  56                  DMH                   RCB
  38                  DMK                   ROM
  1823                DND                   RSK
  1832                DNI                   RWC
  1829                DOS                   RTO
  1825                DPC                   RNO
  18                  DPG                   ROM
  48                  DQC                   RWC
  50                  DQU                   RCB
  1831                DRM                   RKB
  27                  DSC                   RSC
  1902                DSE                   RKB
  1619                DSI                   RWC
  23                  DSQ                   RSC
  1824                DSS                   RSK
  30                  DVA                   ROM

### CoastInteriorOrgUnit

  **ROLLUP\_REGION\_CODE**   **ORG\_UNIT\_CODE**   **ORG\_UNIT\_NAME**            **Coast\_Interior**
  -------------------------- --------------------- ------------------------------ ---------------------
  RCB                        DCC                   Cariboo-Chilcotin              I
  RCB                        DMH                   100 Mile House                 I
  RCB                        DQU                   Quesnel                        I
  RKB                        DRM                   Rocky Mountain                 I
  RKB                        DSE                   Selkirk                        I
  RNO                        DFN                   Fort Nelson                    I
  RNO                        DPC                   Peace                          I
  ROM                        DJA                   Fort St. James                 I
  ROM                        DMK                   Mackenzie                      I
  ROM                        DPG                   Prince George                  I
  ROM                        DVA                   Vanderhoof                     I
  RSC                        DCK                   Chilliwack                     C
  RSC                        DSC                   Sunshine Coast                 C
  RSC                        DSQ                   Sea to Sky                     C
  RSK                        DKM                   Coast Mountains                I
  RSK                        DND                   Nadina                         I
  RSK                        DSS                   Skeena Stikine                 I
  RTO                        DCS                   Cascades                       I
  RTO                        DKA                   Thompson Rivers                I
  RTO                        DOS                   Okanagan Shuswap               I
  RWC                        DCR                   Campbell River                 C
  RWC                        DNI                   North Island - Central Coast   C
  RWC                        DQC                   Haida Gwaii                    C
  RWC                        DSI                   South Island                   C

### SILV\_FUND\_SRC\_CODE

  **SILV\_FUND\_SRCE\_CODE**   **DESCRIPTION**                                  **Category**
  ---------------------------- ------------------------------------------------ --------------
  BCT                          BC Timber Sales                                  Legal
  CBI                          Carbon Offset Investment                         Other
  CL                           Catastrophic Losses                              Operational
  ERP                          Ecosystem Restoration Program                    Other
  FID                          Forest Investment Account - FRPA 108             FIA
  FIL                          Forest Investment Account - licensee             FIA
  FIM                          Forest Investment Account - ministry             FIA
  FIV                          Forest Investment Vote                           FIA
  FMC                          COMFOR                                           Other
  FME                          Forest Worker Development Program                Other
  FMY                          Youth Fund                                       Other
  FR                           FRDA I & II                                      FRDA
  FRP                          FRPA - Application For Relief                    Legal
  FTL                          Forests for Tomorrow Lic Admin                   FFT
  FTM                          Forests for Tomorrow MOF Admin                   FFT
  GA                           Other Agencies or Voluntary Work                 Other
  GAC                          Corrections                                      Other
  GFS                          Forest Stand Management Fund                     Legal
  GJF                          Job Creation, Federal                            Other
  GJJ                          Job Creation, Joint                              Other
  GJM                          Job Creation, MOF                                Other
  IA                           Industrial Appraisal                             Legal
  IIR                          Industrial Incr., Required                       Legal
  IIV                          Industrial Incr., Voluntary                      Incremental
  IO                           Industry Outstanding                             Backlog
  IR                           Industry Royalties                               Backlog
  JOP                          Job Opportunity Program                          Other
  LFP                          Licensee Funded Program                          Other
  M                            Ministry                                         Operational
  O                            Operational                                      Operational
  RBC                          Forest Renewal B.C.                              FRBC
  RBL                          Forest Renewal B.C. - licensee administered      FRBC
  RBM                          Forest Renewal B.C. - ministry administered      FRBC
  S                            Section 88                                       Backlog
  SBF                          BC Timber Sales                                  Legal
  SMF                          South Moresby Forest Replacement Account         Legal
  TSC                          Tree Seed Centre                                 Other
  VCV                          Provincial Contingency Vote (Special Use Only)   Other
  VFH                          Forest Health                                    Other
  VFP                          Current Reforestation                            Operational
  VG                           Ministry Outstanding                             Backlog
  VI                           Provincial Incremental                           Incremental
  VOB                          Ministry Vote, SBEP                              Backlog
  VOI                          Industry Outstanding                             Backlog
  VRT                          Afforestation                                    Other
  XXX                          No Funding Source                                Other

[^1]: Detailed descriptions and downloads of the source layers are
    available from the B.C. Data Catalogue at:
    <http://catalogue.data.gov.bc.ca/dataset>

