---
topic: "land"
title: "Trends in Forest Tree Seed Use in B.C. (1987-2023)"
output: envreportutils.internal::print_ver
editor_options: 
  markdown: 
    wrap: 72
---

```{=html}
<!--
Copyright 2018 Province of British Columbia

Licensed under the Apache License, Version 2.0 (the "License");
you may not use this file except in compliance with the License.
You may obtain a copy of the License at

http://www.apache.org/licenses/LICENSE-2.0

Unless required by applicable law or agreed to in writing, software distributed under the License is distributed on an "AS IS" BASIS,
WITHOUT WARRANTIES OR CONDITIONS OF ANY KIND, either express or implied.
See the License for the specific language governing permissions and limitations under the License.
-->
```

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library(readr) #read in CSV files
library(dplyr) #data prep
library(reshape2) #for restructuring data table
library(bcmaps) #get bc boundary and nrs boundaries, package from GitHub
library(ggplot2)  #for plotting
library(extrafont) #Verdana font
library(envreportutils) #soe theme & svg_px()
library(rphylopic) #for conifer image
library(curl) #required by rphylopic function
library(scales) #for pretty_breaks()
library(rmapshaper) # for intersect and simplify functions
library(sp) # mapping
library(knitr) #for knitting to PDF
library(here)
library(kableExtra)
library(envreportutils.internal)
library(tibble)

read_chunk("tree_seed_use.R")

options(tinytex.verbose = TRUE)

```

```{r pre}
```

Conserving, protecting and managing forest tree genetic resources is the
foundation for economic, social, cultural and ecological goods and
services that flow from British Columbia's forests. Tracking forest tree
seed use from source to planting site is important for assessing B.C.’s
forest genetic resources. This information supports the continuous
improvement of genetic conservation and resource management plans,
strategies and actions, including seed use best management practices
used in reforestation. This indicator investigates the trends in forest
tree seed use in regenerating B.C.'s forests from 1987 to 2023.

-   Maintaining genetic diversity ensures forest ecosystems are healthy,
    diverse, and productive. Genetically diverse forests are more
    resilient to natural disturbances and climate change.

-   The average area of B.C's crown land reforested through planting
    over the past ten years is approximately 195,700 hectares per year
    (2014-2023). This 10-year average is slightly more than the previous
    average of approximately 177,400 hectares per year (2004-2013).

-   Select seed is tree seed that is collected for specific genetic
    traits such as growth, pest resistance, and wood density. Select
    seed is sourced from orchards and natural stands. The proportion of
    crown land planted with select seed has steadily increased over the
    past two decades and was 68% of tree seed use in 2023.

-   Climate Based Seed Transfer is a climate adaptation strategy that
    matches seed sources (seedlots) to climatically suitable planting
    sites. The strategy was fully implemented in 2022 through amendments
    to the [Chief Forester Standards for Seed
    Use](https://www2.gov.bc.ca/gov/content/industry/forestry/managing-our-forest-resources/tree-seed/legislation-standards/chief-forester-s-standards-for-seed-use)

Visit [Tree
Seed](http://www2.gov.bc.ca/gov/content?id=5D1A44C29890457F83E3D55E5B2FFBD8)
to learn more about tree seed use in B.C., and the provincial [Tree
Improvement
Program](http://www2.gov.bc.ca/gov/content?id=6919718586C44A20B14EE9A9AE6C8B7D),
which helps guide reforestation and silviculture investments through
forest genetic research, tree breeding and seed orchard programs. Visit
[Climate Based Seed
Transfer](https://www2.gov.bc.ca/gov/content?id=3B29FA5964254123BE2C482F43E12FA9)
to learn about the science, policy, and tools available to help make
suitable climate based seed transfer choices.

\newpage

## Forest Regeneration by Tree Seed Source in B.C. (1987-2023)

-   After a disturbance by forest fire, forest pest or forest harvest,
    reforestation occurs by natural regeneration or by planting. Seed
    used for planting may be from one or more genetic sources: orchard,
    natural stand superior provenance (superior seed), and natural stand
    non-superior provenance (non-superior seed).

-   Extensive research trials have shown that orchard and natural stand
    seed sources are better for tree growth, pest resistance, and wood
    density.

-   The area of B.C. that is planted instead of naturally regenerated
    has changed from 1987 to 2023. In 2023, the most used seed source
    was natural stand non-superior (45.8%), followed by orchard (45.5%),
    natural regeneration (4.7%), and natural stand superior (3.9%).
    Prior to 2023, orchards were the most common seed source over the
    past 10 years.

```{r bc_regen, fig.height = 6, fig.width = 11}
```

\newpage

## Select Seed Use by Natural Resource District in B.C.

-   Trends in select seed use are variable at the natural resource
    district level. This is partly due to the availability of
    regionally-specific tree species in orchards. Annual or periodic
    fluctuations in orchard cone crops and seed inventories also
    influence select seed use at local levels.

-   Select seed use has increased provincially at a steady rate since
    the early nineties: 1987 - 0%, 1994 - 12%, 2001 - 35%, 2009 - 51%,
    2016 - 64% and 2023 - 52%.

-   The maps below display the percentage of select seed use by natural
    resource district reported in 6-year snapshots from 1987 to 2023. In
    2023, the percentage of select seed used ranged from 15% in Omnica
    Natural Resource District to 94% in Haida Gwaii Natural Resource
    District.

```{r map_regen, fig.height = 6, fig.width = 11}
```

\vspace{20pt}

\newpage

## References and Other Useful Links

-   Learn more about [Tree
    Seed](http://www2.gov.bc.ca/gov/content?id=5D1A44C29890457F83E3D55E5B2FFBD8)
-   Learn about [trends in silviculture in B.C.](silviculture.html)
-   [Operational Tree Improvement
    Program](http://www2.gov.bc.ca/gov/content?id=6919718586C44A20B14EE9A9AE6C8B7D)
-   [Seed Planning and Registry Application
    (SPAR)](http://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)
-   [Chief Forester Standards for Seed
    Use](https://www2.gov.bc.ca/gov/content?id=322E559CA1E44A0294CF656FA0E6EE1E)
-   [Climate Based Seed
    Transfer](https://www2.gov.bc.ca/gov/content?id=3B29FA5964254123BE2C482F43E12FA9)
-   [Forest Genetics Council of British
    Columbia](http://www.fgcouncil.bc.ca/)

## Data

\*By accessing these datasets, you agree to the license associated with
each file, as indicated in parentheses below.

-   [Indicator Summary Data: Forest Regeneration by Tree Seed Source in
    B.C.
    (1987-2019)](https://catalogue.data.gov.bc.ca/dataset/54ec827b-3b9a-4fea-8d9b-d8c006e5b9cc)
    (Licence: [Open Government Licence - British
    Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
-   [Indicator Summary Data: Select Seed Use by Natural Resource
    District in B.C.
    (1987-2019)](https://catalogue.data.gov.bc.ca/dataset/54ec827b-3b9a-4fea-8d9b-d8c006e5b9cc)
    (Licence: [Open Government Licence - British
    Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

For more information on this indicator or on Forest Tree Seed Use in
British Columbia contact the Forest Improvement and Research Management
Branch at [FORHTIP.SEEDHELP\@gov.bc.ca](FORHTIP.SEEDHELP@gov.bc.ca).

Published and Available Online at Environmental Reporting BC (May 2018):
<http://www.env.gov.bc.ca/soe/indicators/land/tree-seed-use.html>

Email correspondence to:
[envreportbc\@gov.bc.ca](mailto:envreportbc@gov.bc.ca){.email}

\newpage

# Glossary

**Climate based seed transfer** is an important climate change
adaptation strategy to promote healthy, resilient and productive forests
and ecosystems through the matching of seedlings/seedlots to future
(projected) planting site climates. In BC, the movement of
seed/seedlings to sites that represent the predicted climate appropriate
for the seed is small, representing only a quarter of a rotation into
the future (i.e., 20 years in the Interior and 15 years on the Coast).
This ‘quarter rotation’ adjustment will be updated as we move forward in
time.

**Genetic diversity** is the genetic variation within individual
organisms, within populations and among populations of a species. A
fundamental component of biological diversity (genetic, species and
ecosystem), genetic diversity plays an important role in the survival,
adaptability and productivity of species and populations.

**Natural stand non-superior provenance seed** is wild seed collected
from a natural stand seed source having no “known” level of genetic
improvement.

**Natural stand superior provenance seed** is wild seed collected from a
natural stand seed source identified as a superior provenance with a
“known” level of genetic improvement for a specific trait (e.g. growth)
over that of local wild stand populations.\

**Orchard seed** is collected from parent trees growing in a managed
seed orchard that is selected and tested as part of a tree breeding
program.

**Select seed** is collected from either orchards or natural stand
superior provenances. Select seed exhibits a higher level of improvement
in one or more desired genetic traits (such as growth, form, wood
density, and resistance to insects and disease) than wild seed collected
from an average natural stand.

**Tree Improvement Program**, in BC is set up to provide strategic and
long term planning, administrative oversight and management of forest
tree seed orchards and tree breeding for the purposes of producing seed
for reforestation that is genetically suitable (adapted to its planting
site) and having a level of improvement (i.e. a known genetic value over
that of “wild” seed based on science-based forest genetic theory and
research).


