---
topic: "land"
title: "Trends in Forest Tree Seed Use in B.C. (1987-2016)"
output: envreportbc::print_ver
---

```{r set-options, echo=FALSE, cache=FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(echo=FALSE, cache=FALSE, message=FALSE, warning=FALSE, 
                      error=FALSE)

library(readr) #read in CSV files
library(dplyr) #data prep
library(reshape2) #for restructuring data table
library(bcmaps) #get bc boundary and nrs boundaries, package from GitHub
library(bcmaps.rdata) # maps data
library(ggplot2)  #for plotting
library(extrafont) #Verdana font
library(envreportutils) #soe theme & svg_px()
library(rphylopic) #for conifer image
library(curl) #required by rphylopic function
library(scales) #for pretty_breaks()
library(rmapshaper) # for intersect and simplify functions
library(sp) # mapping
library(knitr) #for knitting to PDF

read_chunk("tree_seed_use.R")
```

```{r pre}
```

Conserving, protecting and managing forest tree genetic resources is the
foundation for economic, social, cultural and ecological goods and
services that flow from British Columbia\'s forests. Tracking forest
tree seed use---from source to planting site---is an important tool for
monitoring and assessment of B.C.'s forest tree genetic resources. This
information supports the continuous improvement of genetic conservation
and resource management plans, strategies and actions, including seed
use best management practices used in reforestation. This indicator
investigates the trends in forest tree seed use in regenerating B.C.\'s
forests from 1987 to 2016.

-   Maintaining genetic diversity, a fundamental component of
    biodiversity, ensures forests and ecosystems are healthy, resilient
    and productive. It also plays an important role in adaptation so
    that forests and ecosystems are able to effectively respond to
    natural disturbance events---such as wildfire, pests and disease
    infestations---as well as the impacts of climate change.
-   The area of B.C\'s crown land reforested through planting has
    remained consistent---with an average over the past ten years of
    approximately 186,000 hectares per year (2007-2016), a slight
    increase over the previous 10 year average of approximately 170,000
    hectares per year (1997-2006).
-   Select seed---seed selected for specific genetic traits such as
    growth, pest resistance, and wood density---includes seed sourced
    from orchards and natural stand superior provenances. Over the past
    two decades, the proportion of crown land planted with select seed
    has steadily increased to 70 percent in 2016.
    
Visit [Tree Seed](http://www2.gov.bc.ca/gov/content?id=5D1A44C29890457F83E3D55E5B2FFBD8)
to learn more about tree seed use in B.C., and the provincial [Tree
Improvement
Program](http://www2.gov.bc.ca/gov/content?id=6919718586C44A20B14EE9A9AE6C8B7D),
which helps guide reforestation and silviculture investments through
forest genetic research, tree breeding and seed orchard programs.

\newpage

## Forest Regeneration by Tree Seed Source in B.C. (1987--2016)

\vspace{20pt}

```{r bc_regen, fig.height = 6, fig.width = 11}
```

\vspace{20pt}

-   The chart above displays the change in reforestation
    patterns---planted versus natural regeneration---and the change in
    area of forest planted by tree seed source from 1987 to 2016.
-   After a disturbance by forest fire, forest pest or forest harvest,
    reforestation occurs by natural regeneration, meaning from roots or
    seed, or by planting. Seed used for planting may be from one or more
    genetic sources: orchard, natural stand superior provenance, and
    natural stand non-superior provenance.
-   Both orchard and natural stand superior provenance seed sources have
    a known level of improvement based on extensive forest genetic
    adaptation research trials and tree breeding for the selection of
    desired traits---such as growth, pest resistance and wood density.

\newpage

## Select Seed Use by Natural Resource District in B.C.

\vspace{20pt}

```{r map_regen, fig.height = 6, fig.width = 11}
```

\vspace{20pt}

-   The chart above displays the percentage of select seed use by
    natural resource district reported in 5-year snapshots from 1991
    to 2016.
-   Select seed use has increased provincially at a steady rate since
    the early nineties: 1991 -- 3%, 1996 -- 18%, 2001 -- 35%, 2006 --
    50%, 2011 -- 63% and 2016 -- 67%.
-   Select seed use at the natural resource district level is more
    variable, in part due to variances in the number and geographic
    range of tree species for which there are orchards. Annual or
    periodic fluctuations in orchard cone crops and seed inventories
    also influence select seed use at local levels.
-   Overall, the increased use of select seed is indicative of the
    substantial effort underway to meet the target set by the Forest
    Genetic Council of British Columbia to, by 2020, increase select
    seed use to 75 percent of the total provincial seed use for
    reforestation.

\newpage

## References and Other Useful Links

* Learn more about [Tree
    Seed](http://www2.gov.bc.ca/gov/content?id=5D1A44C29890457F83E3D55E5B2FFBD8)
* Learn about [trends in silviculture in B.C.](silviculture.html)
* [Operational Tree Improvement
    Program](http://www2.gov.bc.ca/gov/content?id=6919718586C44A20B14EE9A9AE6C8B7D)
* [Seed Planning and Registry Application
    (SPAR)](http://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)
* [Forest Genetics Council of British
    Columbia](http://www.fgcouncil.bc.ca/)

## Data

\*By accessing these datasets, you agree to the license associated with each file, as indicated in parentheses below.

* [Indicator data: Forest Regeneration by Seed Source in B.C.
    (1987-2016)]() (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))
* [Indicator data: Forest Regeneration by Seed Source by Natural
    Resource District in B.C. (1987-2016)]() (Licence: [Open Government Licence - British Columbia](http://www2.gov.bc.ca/gov/content?id=A519A56BC2BF44E4A008B33FCF527F61))

Published and Available On-Line at Environmental Reporting BC (December 2017):
[http://www.env.gov.bc.ca/soe/indicators/land/tree-seed-use.html](http://www.env.gov.bc.ca/soe/indicators/land/tree-seed-use.html)

Email correspondence to: envreportbc@gov.bc.ca

\newpage

# Glossary

- **Genetic diversity** is the genetic variation within individual organisms, within populations and among populations of a species. A fundamental component of biological diversity (genetic, species and ecosystem), genetic diversity plays an important role in the survival, adaptability and productivity of species and populations.
- **Natural stand non-superior provenance** seed is wild seed collected from a natural stand seed source having no "known"" level of genetic improvement. 
- **Natural stand superior provenance** seed is wild seed collected from a natural stand seed source identified as a superior provenance with a "known" level of genetic improvement for a specific trait (e.g. growth) over that of local wild stand populations. 
- **Orchard** seed is collected from parent trees growing in a managed seed orchard that is selected and tested as part of a tree breeding program.
- **Select** seed is collected from either orchards or natural stand superior provenances. Select seed exhibits a higher level of improvement in one or more desired genetic traits (such as growth, form, wood density, and resistance to insects and disease) than wild seed collected from an average natural stand. 
- **Tree improvement program** in BC is set up to provide strategic and long term planning, administrative oversight and management of forest tree seed orchards and tree breeding for the purposes of producing seed for reforestation that is genetically suitable (adapted to its planting site) and having a level of improvement (i.e. a known genetic value over that of "wild" seed based on science-based forest genetic theory and research).  

\newpage

# METHODS

The [**R** code](http://www.r-project.org) for creating the graphs presented on this page is available on [GitHub](https://github.com/bcgov/EnvReportBC).

## Source Data

The Forest Tree Seed Use analysis uses data from:

1.  [Reporting Silviculture Updates and Land Status Tracking System (RESULTS)](https://www2.gov.bc.ca/gov/content?id=F62992DA5B324446AD5E1B5BFFA404CC>) and

2.  [Seed Planning and Registry System (SPAR)](https://www2.gov.bc.ca/gov/content?id=A2F6834E799241929E4113210A9BDD08)


Wherever possible, the analysis used public data layers from the B.C.
Geographic Warehouse[^1] released under either the Open Government Licence---British Columbia or the B.C. Crown Copyright Licence:

-   **Activity Treatment Unit** –
    WHSE\_FOREST\_VEGATATION\_RSLT\_ACTIVITY\_TREATMENT\_UNIT

    -   <http://catalogue.data.gov.bc.ca/dataset/results-activity-treatment-units>

-   **Opening\_VW** – WHSE\_FOREST\_VEGATATION\_OPENING\_VW

    -   <http://catalogue.data.gov.bc.ca/dataset/results-openings-attribute-only>

-   **Seedlot** – WHSE\_FOREST\_VEGATATION\_SEED\_SEEDLOT

    -   <http://catalogue.data.gov.bc.ca/dataset/seedlot-table>

-   **Veglot** – WHSE\_FOREST\_VEGETATION.SEED\_VEG\_LOT

    -   <http://catalogue.data.gov.bc.ca/dataset/vegetative-lot-table>

-   **Planting** – WHSE\_FOREST\_VEGATATION\_RSLT\_PLANTING\_RSLT

    -   Currently not in the B.C. Geographic Warehouse

-   **Standards Unit** –
    WHSE\_FOREST\_VEGATATION\_RSLT\_STOCKING\_STNDRD\_UNIT

    -   Currently not in the B.C. Geographic Warehouse

-   **Seedlot Genetic Worth** – SEEDLOT\_GENETIC\_WORTH

    -   Currently not in the B.C. Geographic Warehouse

-   **Veglot Genetic Worth** – SEEDLOT\_GENETIC\_WORTH

    -   Currently not in the B.C. Geographic Warehouse

The layers above that are currently not publicly available in the B.C. Geographic Warehouse
are in the process of being added so that the raw data and source code
can be published to allow interested parties to repeat the analysis.

## Planting

RESULTS openings may have one or more planting treatment report that
contains the area treated, species planted, SPAR seedlot/veglot (lot)
and the number of trees planted. To determine the genetic source by
area, the number of trees planted by lot is prorated to the treatment
area and is reported based on the date planted. The reporting for
planted trees by genetic source is based on the planting date.
Classification for the genetic source is:

-   "A": GENETIC\_CLASS\_CODE = "A"

-   "B+": GENETIC\_CLASS\_CODE = "B" and SUPERIOR\_PRVNC\_IND = "Y"

-   "B": GENETIC\_CLASS\_CODE = "B" and SUPERIOR\_PRVNC\_IND = "N"

-   "B" GENETIC\_CLASS\_CODE = "B" and SUPERIOR\_PRVNCE\_IND is null

-   "B" GENETIC\_CLASS\_CODE is null

Note: Reporting of current trends and patterns in reforestation may be
incomplete due to delays in surveying after a disturbance event
(typically a 1-3 year delay in areas planted and a 3-7 year delay in
areas managed for natural regeneration).

## Natural Regeneration

Natural Regeneration is not reported in RESULTS. Instead, this is
inferred by the existence of planting records. If the sum of the
planting area is &lt;= 10% of the opening, the opening is classified as
natural regenerated and the total Net Area to be Reforested (NAR) of the
opening is used as the area of natural regeneration. The year displayed
for natural regeneration is based on the disturbance end date, if not
available, the disturbance start date is used.

Note: Reporting of current trends and patterns in reforestation may be
incomplete due to delays in surveying after a disturbance event
(typically a 1-3 year delay in areas planted and a 3-7 year delay in
areas managed for natural regeneration).

## Fiscal Year

Fiscal year will be display as follows:

-   Formula used: Fiscal Yr:```IIf(Month([DATE])<4,Year([DATE])-1,Year([DATE]))```

-   Example: 1987-04-01 and 1988-03-31 Fiscal Year = 1988

### QUERY STEPS – Baseline Make Table Queries

All queries are built in Microsoft Access 2007. Make Table queries are
used to create baseline data tables for increased efficiency and data
auditability. Note that all data reported after April 1, 1970 is added
to the baseline but the data are subset to more recent years in the
final queries.

### Make Tables – Planting Area: A, B+, B

*1a-Make Planting Activities (outcome: 1a-Planted Activities)*

Joins the Opening\_VW and ATU source tables to create a list of openings
with planting activities completed after March 1970.

  **Field**                   **Criteria**      **Details**
  --------------------------- ----------------- -------------------------------------------------------------
  ATU.ATU\_COMPLETION\_DATE   &gt;=1970-04-01   
  ATU.SILV\_BASE\_CODE        PL                
  ATU.ACT\_PLANTED\_NO        &gt;0             Isolates only those completed activities with planted trees

*1b-Make Planted Trees (outcome: 1b-Planted Trees)*

Joins the Opening, ATU and Planting source tables to create a list of
tree species planted after March 1970.

  **Field**                   **Criteria**      **Details**
  --------------------------- ----------------- -------------------------------------------------------------
  ATU.ATU\_COMPLETION\_DATE   &gt;=1970-04-01   
  ATU.SILV\_BASE\_CODE        PL                
  ATU.ACT\_PLANTED\_NO        &gt;0             Isolates only those completed activities with planted trees

*1c-Make Unique Opening (outcome: 1c-Unique Opening)*

Groups the 1a-Planted Activity table by Opening\_ID to create a unique
list of openings where one or more planting activities have occurred.

*1d-Make Opening (outcome: 1d-Opening)*

Joins the 1c-Unique Opening table with the Opening\_VW and Opening
source tables to add general overview opening information such as
management unit, opening category and disturbance dates. The Opening
source table will not be needed in this query once the Opening\_VW
ORG\_UNIT attribute is correctly referenced – this fix is expected to be
applied to the BC Data Catalogue layers in the summer of 2016.

### QUERY STEPS – Data Summary

### Planting Area: A, B+, B

*2b-Planting Details with Seedlot*

Joins the 1a-Planted Activities, 1b-Planted Trees, 1d-Opening, Org Unit
and Seedlot tables to summarize activity-based details for each
species/seedlot planted.

  -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                             **Criteria**                                                                                                                                                            **Details**
  ------------------------------------  ----------------------------------------------------------------------------------------------------------------------------------------------------------------------- -----------------------------------------------------------------
  Planted Yr                            ```Planted Yr: IIf(Month([ATU_COMPLETION_DATE])<4, Year([ATU_COMPLETION_DATE])-1, Year([ATU_COMPLETION_DATE]))```                                                        Converts date to fiscal year

  SeedClass                             ```A=GENETIC_CLASS_CODE= "A" B+=GENETIC_CLASS_CODE="B" and SUPERIOR_PRVNC_IND="Y" B=GENETIC_CLASS_CODE="B" and SUPERIOR_PRVNCE_IND ="N" B=GENETIC_CLASS_CODE is null```
  
  ProratePLArea                         ```ProratePLArea: Round((([1b-Planted Trees]![NUMBER_PLANTED]/[1b-Planted Trees]![ACT_PLANTED_NO])*[1b-Planted Trees]![TREATMENT_AMOUNT]),2)```                            This gives a prorated area for each of the individual species seedlot planted
  
  
  -----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------

*2c-Summary Seed Class*

Sums the ProratePLArea from the results of query 2b-Planting Details
with Seedlot by Planted Yr and SeedClass for the Province.

*2d-Summary Seed Class BC*

Sums the ProratePLArea from the results of query 2b-Planting Details
with Seedlot by Planted Yr, SeedClass, District and Region.

### Natural Regeneration

*3a-Make Opening Natural Regen (outcome 3a-Opening Natural Regen)*

Joins the Opening\_VW and Opening source tables to provide a unique list
of openings to add general overview opening information such as
management unit, opening category and disturbance dates to be used in
the natural regeneration queries.


 ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
  **Field**                                           **Criteria**                                                                                   **Details**
  --------------------------------------------------  ---------------------------------------------------------------------------------------------  ----------------------------------------   
  Best Date                                           ```IIf([DISTURBANCE_START_DATE] Is Null,[DISTURBANCE_END_DATE], [DISTURBANCE_START_DATE])```   Uses the Disturbance End Date, if it is not available, then use the Disturbance Start Date
  
  OPENING_ VW.OPENING_ GROSS_AREA                      >0.1                                                                                          Excludes shell openings
  
 ------------------------------------------------------------------------------------------------------------------------------------------------------------------------------
 
*3b-Make SU (outcome – 3b-SU)*

Joins the Opening\_VW and SU source tables to summarize the Net Area to
be Reforested (NAR) by OPENING\_ID. NAR is used as a baseline to
calculate the less than 10% planted percentage to be considered a
natural regeneration opening.


  ---------------------------------------------------------------------------
  **Field**                          **Criteria**   **Details**
  ---------------------------------- -------------- -------------------------
  OPENING\_VW.OPENING\_GROSS\_AREA   <>0.1          Excludes shell openings
  
  ---------------------------------------------------------------------------

*3c-PL Area*

Summarizes the 1a-Planted Activities query Planted Area by Opening\_ID
where the Technique is “PL”.


  -----------------------------------------------------------------------
  **Field**               **Criteria**   **Details**
  ----------------------- -------------- --------------------------------
  SILV\_TECHNIQUE\_CODE   ```"PL"```     
  
  TREATMENT\_AREA                        Sums the total PL/PL Area by Opening\_ID

  -----------------------------------------------------------------------

*3d-Combine Opening SU and PL Area*

Joins the 3a-Opening Natural Regen, 3b-SU, 3c-PL Area and Org Unit to
summarize all Planted and NAR areas in one query.

  ---------------------------------------------------------------------------------------------------------
  **Field**         **Criteria**                                                           **Details**
  ----------------  ---------------------------------------------------------------------  ----------------
  Disturb Start Yr  ```IIf(Month([Best Date])<4,Year([Best Date])-1,Year([Best Date]))```  Converts the Best Date to Fiscal Year
  
  ---------------------------------------------------------------------------------------------------------

*3e-Fix Missing NAR and PL Values*

Calculates the 3d-Combine Opening SU and PL Area null values for NAR or
PL Area to zero in order to do subsequent calculations.

  --------------------------------------------------------------------
  **Field**   **Criteria**                               **Details**
  ----------  -----------------------------------------  -------------
  NewNAR      ```iif([NAR] is null,0,[NAR])```
  
  NewPL       ```Iif([PL Area] is null, 0,[PL Area])```   

  ---------------------------------------------------------------------

*3f-Calculate Regen Ratio*

Summarizes the 3e-Fix Missing NAR and PL Values NewNAR by opening. Note
that planting is the sum of the PL/PL. This area is compared against
opening’s NAR. Generally, records with missing NAR are historical
records.

  ---------------------------------------------------------------------------------------------------------------------
  **Field**   **Criteria**                                                           **Details**
  ---------   ---------------------------------------------------------------------  ------------------------------------
  Ratio       ```iif([NewNAR]=0,0,iif([NewNAR]>0, Round([NewPL]/[NewNAR],2),"")```    This defaults 0 value if there is no NAR, otherwise calculate the ratio of planting. Note that records with no NAR will not be accounted for in the total values for natural regen. There are approximately 23K records that is missing NAR
  
  ---------------------------------------------------------------------------------------------------------------------

*3g-Assess Regen Type – Set Threshold*

Uses 3f-Calculate Regen Ratio to define the Regen Type of Natural or
Planted. A threshold of 10% or less of the NAR area planted is defined
as Natural while all openings with &gt; 10% of the NAR area planted are
considered planted.

  -----------------------------------------------------------------------
  **Field**   **Criteria**                                  **Details**
  ----------  --------------------------------------------  -------------
  Regen Type  ```iif(;Ratio]<=0.1,"Natural","Planted")``` 
  
  -----------------------------------------------------------------------

*3h-Final Natural vs Planted BC*

Summarizes the NAR area in 3g-Assess Regen Type-Set by Regen Type. Note
that the Natural Column is what is defined as the total area naturally
regenerated. Planted column is ignored as it is reflected by the
prorated of species trees planted – actuals vs. derived process.

*3i-Final Natural vs Planted Org Unit*

Takes 3g-Assess Regen Type-Set Threshold that has assigned natural and
planted ratio set and groups the sum of NAR area that is by Regen Type.
Note that the Natural Column is what is defined as the total area
naturally regenerated. This data is summarized by Region and District.
Planted column is ignored as it is reflected by the prorated of species
trees planted – actuals vs. derived process.

[^1]: Detailed descriptions and downloads of the source layers are
    available from the B.C. Data Catalogue at:
    <http://catalogue.data.gov.bc.ca/dataset>


